---
title: "Categorical LMTP check"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Categorical LMTP check}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, message=FALSE}
library(modulatR)
```

```{r}
dose_levels <- c(
  "none",
  "low",
  "medium",
  "high"
)

dose_score <- c(
  none = 0,
  low = 1,
  medium = 2,
  high = 3
)
```

The policy is:

If a patient is eligible for treatment escalation, increase treatment by one
category. Otherwise leave treatment unchanged. Patients already receiving
high treatment remain at high treatment.

Thus, among eligible patients,
$none \to low \to medium \to high$. 

## DGP 

```{r}
draw_treatment <- function(n) {
  sample(
    dose_levels,
    size = n,
    replace = TRUE,
    prob = rep(0.25, 4)
  )
}


escalate_one_level <- function(A,
                               eligible) {
  next_level <- c(
    none = "low",
    low = "medium",
    medium = "high",
    high = "high"
  )

  out <- A

  out[eligible == 1] <- unname(
    next_level[
      A[eligible == 1]
    ]
  )

  out
}


generate_observed_data <- function(n) {
  W <- rnorm(n)

  eligible <- as.integer(
    W > -0.25
  )

  L1 <- 0.5 * W +
    0.4 * eligible +
    rnorm(n)

  A1 <- draw_treatment(n)

  L2 <- 0.4 * W +
    0.5 * L1 +
    0.6 * dose_score[A1] +
    rnorm(n)

  A2 <- draw_treatment(n)

  Y <- 1 +
    0.5 * W +
    0.4 * L1 +
    0.7 * L2 -
    0.6 * dose_score[A1] -
    1.0 * dose_score[A2] +
    rnorm(n)

  data.frame(
    W = W,
    eligible = eligible,
    L1 = L1,
    A1 = A1,
    L2 = L2,
    A2 = A2,
    Y = Y,
    stringsAsFactors = FALSE
  )
}
```

## Categorical LMTP Definition 

```{r}
mtp_escalate <- mtp_discrete(
  map_fun = function(A, H_df) {
    escalate_one_level(
      A = A,
      eligible = H_df$eligible
    )
  },
  support = dose_levels,
  name = "escalate_if_eligible"
)

policy_escalate <- repeat_policy_over_time(
  mtp = mtp_escalate,
  tau = 2,
  name = "escalate_if_eligible_sequence"
)
```

Verify the policy works.

```{r}
policy_example <- data.frame(
  eligible = rep(c(0, 1), each = 4),
  A = rep(dose_levels, times = 2),
  stringsAsFactors = FALSE
)

policy_example$A_star <- mtp_escalate$apply_policy(
  A_vec = policy_example$A,
  H_df = policy_example["eligible"]
)

policy_example
```


## Monte Carlo Truth

```{r}
generate_policy_outcome <- function(n) {
  W <- rnorm(n)

  eligible <- as.integer(
    W > -0.25
  )

  L1 <- 0.5 * W +
    0.4 * eligible +
    rnorm(n)

  # Natural treatment at time 1
  A1 <- draw_treatment(n)

  # Modified treatment at time 1
  A1_d <- escalate_one_level(
    A = A1,
    eligible = eligible
  )

  # L2 evolves under the modified treatment history
  L2_d <- 0.4 * W +
    0.5 * L1 +
    0.6 * dose_score[A1_d] +
    rnorm(n)

  # Natural treatment at time 2 under the modified history
  A2 <- draw_treatment(n)

  # Modified treatment at time 2
  A2_d <- escalate_one_level(
    A = A2,
    eligible = eligible
  )

  # Conditional mean of the policy outcome.
  #
  # We omit an additional mean-zero outcome error because it does not affect
  # E[Y^d] and only adds Monte Carlo noise.
  Y_d_mean <- 1 +
    0.5 * W +
    0.4 * L1 +
    0.7 * L2_d -
    0.6 * dose_score[A1_d] -
    1.0 * dose_score[A2_d]

  Y_d_mean
}


set.seed(1001)

n_truth <- 500000

truth <- mean(
  generate_policy_outcome(
    n = n_truth
  )
)

truth
```

## Fit the categorical LMTP

```{r}
set.seed(2002)

n <- 10000

dat <- generate_observed_data(
  n = n
)

ds <- LMTPData$new(
  data = dat,
  W_cols = c(
    "W",
    "eligible"
  ),
  L_cols = list(
    "L1",
    "L2"
  ),
  A_cols = c(
    "A1",
    "A2"
  ),
  Y_col = "Y"
)

ds
```

For the sake of this vignette, which is just a sanity check, we use the 
'oracle' (known) treatment probability mass function in fitting the models. 

```{r}
oracle_g_learner <- function(data) {
  function(newdata) {
    rep(
      0.25,
      newdata$n
    )
  }
}

g_learners <- repeat_lmtp_learner(
  learner_fun = oracle_g_learner,
  tau = ds$tau()
)
```

```{r}
m_learners <- make_glm_m_learners(
  tau = ds$tau(),
  family = stats::gaussian()
)

nuisance_factory <- LMTPNuisanceFactory$new(
  policy_seq = policy_escalate,
  m_learners = m_learners,
  g_learners = g_learners,
  truncate_ratio = c(
    1e-6,
    1e6
  )
)

fluctuation <- LMTPFluctuationSubmodel$new(
  family = stats::gaussian()
)
```

```{r}
fit <- run_tmle_for_LMTP(
  ds = ds,
  nuisance_factory = nuisance_factory,
  fluctuation = fluctuation
)

fit
```

## Comparison with the truth

```{r}
comparison <- data.frame(
  truth = truth,
  estimate = unname(fit$estimate),
  se = unname(fit$se),
  lower = unname(fit$ci["lower"]),
  upper = unname(fit$ci["upper"])
)

comparison$error <- with(
  comparison,
  estimate - truth
)

comparison
```



