---
title: "Conditional LMTPs for Subgroup Specific Effects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Conditional LMTPs for Subgroup Specific Effects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  warning = FALSE,
  message = FALSE
)

set.seed(1)
```

# Conditional LMTPs for Subgroup Specific Effects

This vignette illustrates why it can be inefficient to estimate subgroup-specific
longitudinal modified treatment policy effects by simply stratifying the data.

# Packages

```{r}
library(modulatR)
library(nadir)
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
```

# Data-generating process

We simulate a two-timepoint longitudinal data structure with baseline covariate
\(W\), time-varying covariates \(L_1,L_2\), treatments \(A_1,A_2\), and outcome
\(Y\). The subgroup is rare by construction.

The treatment effect is allowed to vary with \(W\), so the rare subgroup has a
different average effect than the full population.

```{r}

generate_lmtp_data <- function(n, seed = NULL) {
  if (!is.null(seed)) set.seed(seed)

  W <- rnorm(n)
  quantiles <- c(-Inf, quantile(W, seq(0, 1, length.out = 6))[2:5], Inf)
  G <- cut(W, quantiles, include.lowest=TRUE)

  eL1 <- rnorm(n)
  eA1 <- rnorm(n)
  eL2 <- rnorm(n)
  eA2 <- rnorm(n)
  eY  <- rnorm(n)

  L1 <- 0.5 * W + eL1

  A1 <- 0.4 * W + 0.7 * L1 + eA1

  L2 <- 0.3 * W + 0.5 * L1 + 0.7 * A1 + eL2

  A2 <- 0.3 * W + 0.4 * L2 + eA2

  beta1 <- 0.8 + 0.4 * W
  beta2 <- 0.6 + 0.2 * W

  Y <- 1 +
    0.5 * W +
    0.3 * L1 +
    0.5 * L2 +
    beta1 * A1 +
    beta2 * A2 +
    eY

  data.frame(
    id = seq_len(n),
    W = W,
    G = G,
    L1 = L1,
    A1 = A1,
    L2 = L2,
    A2 = A2,
    Y = Y
  )
}
```

# Monte Carlo truth

Because this is a simulation, we can approximate the true subgroup specific 
mean outcomes by simulating a very large population and applying the 
structural equations under the shifted policy.

Under the policy \(A_1^d = A_1 + 1\), the downstream covariate \(L_2\), treatment
\(A_2\), and final outcome are recomputed under the same exogenous errors.

```{r}
compute_truth <- function(n_truth = 5e5, delta = 1, seed = 1) {
  set.seed(seed)

  W <- rnorm(n_truth)
  quantiles <- c(-Inf, quantile(W, seq(0, 1, length.out = 6))[2:5], Inf)
  G <- cut(W, quantiles, include.lowest=TRUE)
  

  eL1 <- rnorm(n_truth)
  eA1 <- rnorm(n_truth)
  eL2 <- rnorm(n_truth)
  eA2 <- rnorm(n_truth)
  eY  <- rnorm(n_truth)

  L1 <- 0.5 * W + eL1

  A1 <- 0.4 * W + 0.7 * L1 + eA1
  L2 <- 0.3 * W + 0.5 * L1 + 0.7 * A1 + eL2
  A2 <- 0.3 * W + 0.4 * L2 + eA2

  beta1 <- 0.8 + 0.4 * W
  beta2 <- 0.6 + 0.2 * W

  Y <- 1 +
    0.5 * W +
    0.3 * L1 +
    0.5 * L2 +
    beta1 * A1 +
    beta2 * A2 +
    eY

  # Counterfactual shifted trajectory.
  A1d <- A1 + delta
  L2d <- 0.3 * W + 0.5 * L1 + 0.7 * A1d + eL2
  A2d <- 0.3 * W + 0.4 * L2d + eA2 + delta

  Yd <- 1 +
    0.5 * W +
    0.3 * L1 +
    0.5 * L2d +
    beta1 * A1d +
    beta2 * A2d +
    eY

  df <- tibble::tibble(G, Y, Yd) 
  summary_df <- df |> group_by(G) |> summarize(EY = mean(Y), EYd = mean(Yd),
                                               EYd_minus_EY = EYd - EY)
  
  summary_df
}

subgroup_truths <- compute_truth()
subgroup_truths
```

# Helper functions

We create a small set of helpers to build an `LMTPData` object, define the
additive policy, construct nuisance learners, and run the two estimators.

The conditional estimator uses the full sample. The naive stratified estimator
first filters to the rare subgroup and then refits the LMTP analysis using only
that subset.

```{r}
make_ds <- function(df) {
  LMTPData$new(
    data = df,
    id_col = "id",
    A_cols = c("A1", "A2"),
    L_cols = list("L1", "L2"),
    W_cols = c("W", "G"),
    Y_col = "Y"
  )
}

make_shift_policy <- function(tau, delta = 1) {
  repeat_policy_over_time(
    mtp_additive_shift(delta = delta),
    tau = tau
  )
}

make_nuisance_factory <- function(ds, delta = 1) {
  policy_shift <- make_shift_policy(tau = ds$tau(), delta = delta)

  m_learners <- make_glm_m_learners(
    tau = ds$tau(),
    family = stats::gaussian()
  )

  g_learners <- make_nadir_density_g_learners(
    tau = ds$tau(),
    learner = nadir::lnr_glm_density
  )
  
  lambda_learners <- make_glm_lambda_learners(
    tau = ds$tau(),
    family = binomial(link='logit'))

  LMTPNuisanceFactory$new(
    policy_seq = policy_shift,
    m_learners = m_learners,
    # g_learners = g_learners,
    lambda_learners = lambda_learners,
    truncate_density = 1e-8,
    truncate_ratio = c(1e-4, 1e4)
  )
}

subgroups <- list(
  `G_eq_1` = function(df) as.integer(df$G) == 1,
  `G_eq_2` = function(df) as.integer(df$G) == 2,
  `G_eq_3` = function(df) as.integer(df$G) == 3,
  `G_eq_4` = function(df) as.integer(df$G) == 4,
  `G_eq_5` = function(df) as.integer(df$G) == 5
)

fluctuation <- LMTPFluctuationSubmodel$new(
  family = stats::gaussian()
)
```

# Conditional LMTP estimator

This estimator fits the longitudinal nuisance functions on the full sample, then
summarizes the doubly robust pseudo-outcome in the rare subgroup.

```{r}
estimate_conditional_lmtp <- function(df, truth, delta = 1) {
  ds <- make_ds(df)
  nf <- make_nuisance_factory(ds, delta = delta)

  fit <- run_subgroup_tmle_for_LMTP(
    ds = ds,
    nuisance_factory = nf,
    fluctuation = fluctuation,
    subgroup_funs = subgroups,
    alpha = 0.05
  )

  out <- 
    tibble::tibble(
      method = "Conditional LMTP",
      estimate = fit$estimate,
      se = fit$se,
      lower = estimate + qnorm(0.025) * se,
      upper = estimate + qnorm(0.975) * se,
      G = as.integer(stringr::str_extract(fit$subgroup_names, pattern = "[0-9]+"))
    )
  
  out
}
```

# Naive stratified estimator

The naive estimator filters to the rare subgroup first, then fits the same
conditional LMTP machinery on the smaller data set. This is intentionally
wasteful: it estimates all longitudinal nuisance functions using only the
subgroup observations.

```{r}
estimate_naive_stratified_lmtp <- function(df, truth, delta = 1) {
  results <- list()
  for (g in 1:5) {
    
    df_g <- df |>
      filter(as.integer(G) == g) |>
      select(-G) |> 
      mutate(id = seq_len(n()))
    
    stratified_estimation <- function(df_g) {
      
      ds_g <- LMTPData$new(
        data = df_g,
        id_col = "id",
        A_cols = c("A1", "A2"),
        L_cols = list("L1", "L2"),
        W_cols = c("W"),
        Y_col = "Y"
      )
      
      nf_g <- make_nuisance_factory(ds_g, delta = delta)
      
      fit_g <- run_tmle_for_LMTP(ds = ds_g, 
        nuisance_factory = nf_g,
        fluctuation = fluctuation
      )
      
      out <- list(
        method = 'Naive Stratified LMTP',
        estimate = fit_g$estimate,
        se = fit_g$se,
        lower = fit_g$ci[1],
        upper = fit_g$ci[2])
      
      out
    }
  
    results_g <- stratified_estimation(df_g)
    
    results_g$G <- g
    
    results[[length(results)+1]] <- results_g
  }
  results <- bind_rows(results)
  results
}
```

# One simulation replicate

```{r}
one_rep <- function(n, rep_id, truth, delta = 1) {
  df <- generate_lmtp_data(
    n = n,
    seed = 100000 + 1000 * n + rep_id
  )
  
  subgroup_truths$G <- as.integer(subgroup_truths$G)
  subgroup_truths_for_join <- subgroup_truths |> select(G, EYd)

  bind_rows(
    estimate_conditional_lmtp(df, truth = truth, delta = delta),
    estimate_naive_stratified_lmtp(df, truth = truth, delta = delta)
  ) |>
    mutate(
      n = n,
      rep_id = rep_id
    ) |> 
    left_join(subgroup_truths_for_join, by = 'G') |> 
    mutate(covers = lower <= EYd & estimate <= upper)
}

one_rep(n = 500, rep_id = 1, truth = subgroup_truths)
```

# Simulation across sample sizes

For a real simulation study, increase `B` to 500 or 1000. For a small vignette,
we keep `B` modest.

```{r}
sample_sizes <- (seq(8, 30, length.out=6))^2
B <- 250

sim_results <- tidyr::crossing(
  n = sample_sizes,
  rep_id = seq_len(B)
) |>
  mutate(result = purrr::map2(
    n,
    rep_id,
    ~ one_rep(n = .x, rep_id = .y, truth = subgroup_truths)
  )) |>
  select(result) |>
  tidyr::unnest(result)

# head(sim_results)
```

# Operating characteristics

We summarize bias, root mean squared error, empirical coverage, power against
the null contrast \(0\), average estimated standard error, and average confidence
interval width.

```{r}
sim_summary <- sim_results |>
  filter(!is.na(estimate)) |>
  group_by(n, method, G) |>
  summarize(
    bias = mean(estimate - EYd),
    rmse = sqrt(mean((estimate - EYd)^2)),
    empirical_sd = sd(estimate),
    mean_se = mean(se),
    coverage = mean(covers),
    mean_ci_width = mean(abs(upper - lower)),
    .groups = "drop"
  )

sim_summary
```

# Coverage

The main expectation is that the conditional LMTP approach should have more
stable finite-sample behavior when the subgroup is rare. It estimates the
longitudinal nuisance functions using all available observations, whereas the
naive stratified approach must learn those same nuisance functions from only
about 10% of the original sample.

```{r}
ggplot(sim_summary, aes(x = n, y = coverage, group = method, linetype = method)) +
  geom_hline(yintercept = 0.95, linewidth = 0.4) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = sample_sizes) +
  coord_cartesian(ylim = c(0, 1)) +
  facet_wrap(~G) + 
  labs(
    x = "Total sample size",
    y = "Empirical 95% coverage",
    linetype = NULL,
    title = "Subgroup coverage across sample sizes",
    subtitle = "The naive stratified approach estimates nuisance functions using only each subgroup one at a time."
  ) +
  theme_minimal(base_size = 12) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Power

Because the true subgroup contrast is positive, we can also compare the
probability that the 95% confidence interval excludes zero.

```{r}
# ggplot(sim_summary, aes(x = n, y = power, group = method, linetype = method)) +
#   geom_line(linewidth = 0.8) +
#   geom_point(size = 2) +
#   scale_x_continuous(breaks = sample_sizes) +
#   coord_cartesian(ylim = c(0, 1)) +
#   labs(
#     x = "Total sample size",
#     y = "Power: 95% CI excludes zero",
#     linetype = NULL,
#     title = "Rare-subgroup power across sample sizes"
#   ) +
#   theme_minimal(base_size = 12)
```

# Precision

The confidence intervals for the naive stratified approach are expected to be
wider, especially at small sample sizes, because the effective sample size is the
rare subgroup size.

```{r}
ggplot(sim_summary, aes(x = n, y = mean_ci_width, group = method, linetype = method)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = sample_sizes) +
  labs(
    x = "Total sample size",
    y = "Mean 95% CI width",
    linetype = NULL,
    title = "Confidence interval width for subgroup LMTP contrast"
  ) +
  theme_minimal(base_size = 12) + 
  facet_wrap(~ G)
```

# RMSE

```{r}
ggplot(sim_summary, aes(x = n, y = rmse, group = method, linetype = method)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = sample_sizes) +
  labs(
    x = "Total sample size",
    y = "Root mean squared error",
    linetype = NULL,
    title = "RMSE for subgroup LMTP contrast"
  ) +
  theme_minimal(base_size = 12)
```
