---
title: "Using cross-fitted nuisance learners"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using cross-fitted nuisance learners}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)

if (!requireNamespace("origami", quietly = TRUE)) {
  knitr::knit_exit()
}

set.seed(2026)

library(modulatR)
library(nadir)
library(origami)
```

The nuisance learners in `{modulatR}` have a simple contract: 
each learner takes one `LMTPData` object and returns a prediction function. 
Cross-fitting is implemented as a wrapper around such learners. 
This means that the nuisance factory does not need to know whether a learner is cross-fitted.
(In fact, nuisance factories can be completely blind/agnostic to the internals
of what nuisance fitting function code does -- it just needs to accept data, and
return a prediction function. To work correctly, such nuisance function fitting 
code -- referred to as `m_learners`, `g_learners` or `r_learners` in the code -- 
will need to be correctly specified). 

```{r}
# Data-generating process --------------------------------------------------

set.seed(1)

n <- 1500

W  <- rnorm(n)
L1 <- 0.5 * W + rnorm(n)
A1 <- 0.6 * W + 0.8 * L1 + rnorm(n)
L2 <- 0.4 * W + 0.7 * A1 + 0.6 * L1 + rnorm(n)
A2 <- 0.3 * W + 0.5 * L2 + rnorm(n)

Y <- 1 + 0.5 * W + 0.4 * L1 + 0.3 * L2 + 1.2 * A1 + 0.8 * A2 + rnorm(n)

df <- data.frame(W, L1, A1, L2, A2, Y)

ds <- LMTPData$new(
  data = df,
  A_cols = c("A1", "A2"),
  L_cols = list("L1", "L2"),
  W_cols = "W",
  Y_col = "Y"
)

ds
```

# Policy Definition

We use a simple additive-shift LMTP that increases each exposure by one unit.

```{r}
policy_shift <- repeat_policy_over_time(
  mtp_additive_shift(delta = 1),
  tau = ds$tau()
)

policy_shift
```

# Check the crossfitting  of a learner manually

This is sort of an overly simple example, but meant
to illustrate the point of crossfitting a learner. 

```{r}
base_m_learner <- make_glm_m_learner(
  family = stats::gaussian()
)

cf_m_learner <- crossfit_lmtp_learner(
  learner = base_m_learner,
  V = 3
)

# Prepare a time-indexed dataset with a pseudo-outcome.
ds_t <- ds |>
  modulatR:::.with_policy_sequence(policy_shift) |>
  modulatR:::.with_lmtp_time(2) |>
  modulatR:::.with_pseudo_outcome(ds$Y())

m_predict <- cf_m_learner(ds_t)

is.function(m_predict)

m_pred <- m_predict(ds_t)

length(m_pred)
head(m_pred)
summary(m_pred)
```

The output should show that `m_predict` is a function and that it returns one prediction per row of `ds_t`.

# Fit with cross-fit nuisance learners

```{r}
m_learners <- lapply(
  make_glm_m_learners(
    tau = ds$tau(),
    family = stats::gaussian()
  ),
  crossfit_lmtp_learner,
  V = 3
)

g_learners <- lapply(
  make_nadir_density_g_learners(
    tau = ds$tau(),
    learner = nadir::lnr_glm_density
  ),
  crossfit_lmtp_learner,
  V = 3
)
```

# Specify a Nuisance Factory 

```{r}
nf_shift_cf <- LMTPNuisanceFactory$new(
  policy_seq = policy_shift,
  m_learners = m_learners,
  g_learners = g_learners,
  truncate_density = 1e-12,
  truncate_ratio = c(1e-6, Inf)
)

fluctuation <- LMTPFluctuationSubmodel$new(
  family = stats::gaussian(),
  use_intercept = TRUE
)
```

# Fit a TMLE with Crossfit Nuisances 

```{r}
fit_shift_cf <- run_tmle_for_LMTP(
  ds = ds,
  nuisance_factory = nf_shift_cf,
  fluctuation = fluctuation
)

fit_shift_cf
```


# Now an example where crossfitting is more useful

```{r}
# A nonlinear example where cross-fitting is more relevant -----------------

# In the previous example, the data-generating process is linear and the GLM
# nuisance learners are essentially correctly specified. Here we create a more
# nonlinear DGP and use a ranger learner for the outcome regressions. This makes
# the use of cross-fitting less cosmetic.

# Truth by large Monte Carlo sample ----------------------------------------

set.seed(101)

n_truth <- 2e5

W   <- rnorm(n_truth)
eL1 <- rnorm(n_truth)
eA1 <- rnorm(n_truth)
eL2 <- rnorm(n_truth)
eA2 <- rnorm(n_truth)
eY  <- rnorm(n_truth)

L1 <- 0.5 * W + 0.3 * sin(W) + eL1
A1 <- 0.6 * W + 0.8 * L1 + 0.4 * W^2 + eA1
L2 <- 0.4 * W + 0.7 * A1 + 0.6 * L1 + 0.5 * sin(A1) + eL2
A2 <- 0.3 * W + 0.5 * L2 + 0.25 * L1^2 + eA2

Y <- 1 +
  0.5 * W +
  0.4 * L1 +
  0.3 * L2 +
  1.2 * A1 +
  0.8 * A2 +
  0.7 * sin(A1) +
  0.5 * cos(L2) +
  0.25 * A1 * L1 +
  eY

A1d <- A1 + 1
L2d <- 0.4 * W + 0.7 * A1d + 0.6 * L1 + 0.5 * sin(A1d) + eL2
A2d <- 0.3 * W + 0.5 * L2d + 0.25 * L1^2 + eA2 + 1

Yd <- 1 +
  0.5 * W +
  0.4 * L1 +
  0.3 * L2d +
  1.2 * A1d +
  0.8 * A2d +
  0.7 * sin(A1d) +
  0.5 * cos(L2d) +
  0.25 * A1d * L1 +
  eY

truth_nonlinear <- c(
  EY = mean(Y),
  EYd = mean(Yd),
  diff = mean(Yd - Y)
)

truth_nonlinear

# Observed nonlinear sample ------------------------------------------------

set.seed(102)

n_nonlinear <- 3000

W  <- rnorm(n_nonlinear)
L1 <- 0.5 * W + 0.3 * sin(W) + rnorm(n_nonlinear)
A1 <- 0.6 * W + 0.8 * L1 + 0.4 * W^2 + rnorm(n_nonlinear)
L2 <- 0.4 * W + 0.7 * A1 + 0.6 * L1 + 0.5 * sin(A1) + rnorm(n_nonlinear)
A2 <- 0.3 * W + 0.5 * L2 + 0.25 * L1^2 + rnorm(n_nonlinear)

Y <- 1 +
  0.5 * W +
  0.4 * L1 +
  0.3 * L2 +
  1.2 * A1 +
  0.8 * A2 +
  0.7 * sin(A1) +
  0.5 * cos(L2) +
  0.25 * A1 * L1 +
  rnorm(n_nonlinear)

df_nonlinear <- data.frame(W, L1, A1, L2, A2, Y)

ds_nonlinear <- LMTPData$new(
  data = df_nonlinear,
  A_cols = c("A1", "A2"),
  L_cols = list("L1", "L2"),
  W_cols = "W",
  Y_col = "Y"
)

policy_shift_nonlinear <- repeat_policy_over_time(
  mtp_additive_shift(delta = 1),
  tau = ds_nonlinear$tau()
)

# Flexible m learners using ranger ----------------------------------------

# This assumes nadir::lnr_ranger follows the same learner contract used by
# nadir learners in the rest of the package. If your local nadir::lnr_ranger
# takes ranger-specific arguments, add them here.
m_learners_ranger <- repeat_lmtp_learner(
  make_nadir_m_learner(
    learner = nadir::lnr_ranger
  ),
  tau = ds_nonlinear$tau()
)

m_learners_ranger_cf <- lapply(
  m_learners_ranger,
  crossfit_lmtp_learner,
  V = 3
)

# Keep the treatment density model simple for this example. The main point is
# to show the cross-fitting wrapper sitting on top of a flexible outcome learner.
g_learners_density <- make_nadir_density_g_learners(
  tau = ds_nonlinear$tau(),
  learner = nadir::lnr_glm_density
)

g_learners_density_cf <- lapply(
  g_learners_density,
  crossfit_lmtp_learner,
  V = 3
)

# Plain flexible fit -------------------------------------------------------

nf_shift_ranger <- LMTPNuisanceFactory$new(
  policy_seq = policy_shift_nonlinear,
  m_learners = m_learners_ranger,
  g_learners = g_learners_density,
  truncate_density = 1e-12,
  truncate_ratio = c(1e-6, Inf)
)

fit_shift_ranger <- run_tmle_for_LMTP(
  ds = ds_nonlinear,
  nuisance_factory = nf_shift_ranger,
  fluctuation = fluctuation
)

# Cross-fitted flexible fit ------------------------------------------------

nf_shift_ranger_cf <- LMTPNuisanceFactory$new(
  policy_seq = policy_shift_nonlinear,
  m_learners = m_learners_ranger_cf,
  g_learners = g_learners_density_cf,
  truncate_density = 1e-12,
  truncate_ratio = c(1e-6, Inf)
)

fit_shift_ranger_cf <- run_tmle_for_LMTP(
  ds = ds_nonlinear,
  nuisance_factory = nf_shift_ranger_cf,
  fluctuation = fluctuation
)

# Compare against truth ----------------------------------------------------

comparison_nonlinear <- data.frame(
  estimator = c("truth", "ranger", "cross-fitted ranger"),
  estimate = c(
    truth_nonlinear["EYd"],
    fit_shift_ranger$estimate,
    fit_shift_ranger_cf$estimate
  ),
  se = c(
    NA_real_,
    fit_shift_ranger$se,
    fit_shift_ranger_cf$se
  ),
  error = c(
    0,
    fit_shift_ranger$estimate - truth_nonlinear["EYd"],
    fit_shift_ranger_cf$estimate - truth_nonlinear["EYd"]
  )
)

comparison_nonlinear

fit_shift_ranger
fit_shift_ranger_cf
```

