---
title: "DR-Learner for conditional LMTP effects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DR-Learner for conditional LMTP effects}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)

set.seed(2026)

library(modulatR)
library(nadir)
library(ggplot2)
```

# DR-Learner for conditional LMTP effects

The subgroup TMLE estimates a finite vector of subgroup-specific LMTP parameters. That is useful when the scientific question is about a small number of pre-specified groups.

The DR-learner approach targets a different object: a conditional LMTP function. Here we estimate

\[
\tau(v)
=
\mathbb E\{Y(\bar A^d) - Y \mid W = v\}.
\]

This is useful when the conditioning variable is continuous and we want to recover a smooth effect curve.

# Truth by large Monte Carlo sample

We generate a longitudinal data structure with two time points. The modified treatment policy shifts both treatments upward by one unit:

\[
A_t^d = A_t + 1.
\]

The outcome model is constructed so that the effect of this shift varies smoothly with baseline \(W\).

```{r}
# Truth by large Monte Carlo sample ----------------------------------------

set.seed(1)

n_truth <- 5e5

W   <- rnorm(n_truth)
eL1 <- rnorm(n_truth)
eA1 <- rnorm(n_truth)
eL2 <- rnorm(n_truth)
eA2 <- rnorm(n_truth)
eY  <- rnorm(n_truth)

L1 <- 0.5 * W + eL1

A1 <- 0.5 * W +
  0.7 * L1 +
  0.2 * W^2 +
  eA1

L2 <- 0.4 * W +
  0.6 * L1 +
  0.6 * A1 +
  eL2

A2 <- 0.3 * W +
  0.6 * L2 +
  0.15 * W^2 +
  eA2

# The treatment coefficients vary with W. This makes the conditional
# LMTP contrast a smooth nonlinear function of W.
beta_A1 <- 0.6 + 0.4 * W + 0.25 * W^2
beta_A2 <- 0.5 - 0.25 * W + 0.15 * W^2

Y <- 1 +
  0.5 * W +
  0.3 * L1 +
  0.25 * L2 +
  beta_A1 * A1 +
  beta_A2 * A2 +
  eY

# Counterfactual world under additive shift.
A1d <- A1 + 1

L2d <- 0.4 * W +
  0.6 * L1 +
  0.6 * A1d +
  eL2

A2d <- 0.3 * W +
  0.6 * L2d +
  0.15 * W^2 +
  eA2 +
  1

Yd <- 1 +
  0.5 * W +
  0.3 * L1 +
  0.25 * L2d +
  beta_A1 * A1d +
  beta_A2 * A2d +
  eY

truth_df <- data.frame(
  W = W,
  Y = Y,
  Yd = Yd,
  diff = Yd - Y
)

# Approximate the true conditional effect curve by binning W finely.
truth_curve <- truth_df |>
  transform(W_bin = cut(W, breaks = seq(-2.5, 2.5, length.out = 81))) |>
  subset(!is.na(W_bin)) |>
  aggregate(diff ~ W_bin, data = _, FUN = mean)

truth_curve$W_mid <- sapply(
  strsplit(gsub("\\(|\\]|\\[|\\)", "", as.character(truth_curve$W_bin)), ","),
  function(x) mean(as.numeric(x))
)

truth_curve <- truth_curve[, c("W_mid", "diff")]
names(truth_curve) <- c("W", "truth")

head(truth_curve)
```

# Observed data

```{r}
# Observed data ------------------------------------------------------------

set.seed(2)

n <- 5000

W  <- rnorm(n)
L1 <- 0.5 * W + rnorm(n)

A1 <- 0.5 * W +
  0.7 * L1 +
  0.2 * W^2 +
  rnorm(n)

L2 <- 0.4 * W +
  0.6 * L1 +
  0.6 * A1 +
  rnorm(n)

A2 <- 0.3 * W +
  0.6 * L2 +
  0.15 * W^2 +
  rnorm(n)

beta_A1 <- 0.6 + 0.4 * W + 0.25 * W^2
beta_A2 <- 0.5 - 0.25 * W + 0.15 * W^2

Y <- 1 +
  0.5 * W +
  0.3 * L1 +
  0.25 * L2 +
  beta_A1 * A1 +
  beta_A2 * A2 +
  rnorm(n)

df <- data.frame(W, L1, A1, L2, A2, Y)

ds <- LMTPData$new(
  data = df,
  A_cols = c("A1", "A2"),
  L_cols = list("L1", "L2"),
  W_cols = "W",
  Y_col = "Y"
)

ds
```

# Define the LMTP

```{r}
# Policy ------------------------------------------------------------------

policy_shift <- repeat_policy_over_time(
  mtp_additive_shift(delta = 1),
  tau = ds$tau()
)

policy_shift
```

# Specify nuisance learners

For the first-stage nuisance functions, we use flexible outcome regressions and density learners. In a serious analysis, one would usually cross-fit these nuisance learners. This vignette keeps the computation modest.

```{r}
# Nuisance learners --------------------------------------------------------

m_formula_t1 <- `..pseudo_outcome` ~ W + I(W^2) + L1 + I(L1^2) + A1 + I(A1^2) + W:A1
m_formula_t2 <- `..pseudo_outcome` ~ W + I(W^2) + L1 + I(L1^2) + A1 +
  L2 + I(L2^2) + A2 + I(A2^2) + W:A2 + L2:A2

m_learners <- make_glm_m_learners(
  tau = ds$tau(),
  family = stats::gaussian(),
  formula = list(m_formula_t1, m_formula_t2)
)

g_learners <- make_nadir_density_g_learners(
  tau = ds$tau(),
  learner = nadir::lnr_glm_density
)

nf_shift <- LMTPNuisanceFactory$new(
  policy_seq = policy_shift,
  m_learners = m_learners,
  g_learners = g_learners,
  truncate_density = 1e-12,
  truncate_ratio = c(1e-6, 1e6)
)

fluctuation <- LMTPFluctuationSubmodel$new(
  family = stats::gaussian(),
  use_intercept = TRUE
)
```

# Fit the DR-learner

The meta-learner regresses the DR pseudo-outcome contrast on \(W\). We use a flexible learner because the true curve is nonlinear.

```{r}
# DR-learner ---------------------------------------------------------------

dr_fit <- run_drlearner_for_subgroup_LMTP(
  ds = ds,
  nuisance_factory = nf_shift,
  fluctuation = fluctuation,
  metalearner = make_nadir_metalearner(
    learner = nadir::lnr_hal,
    smoothness_orders = 1
  ),
  conditioning_cols = "W",
  A_type = "continuous"
)

dr_fit
```

# Predict the conditional LMTP contrast

```{r}
# Prediction grid ----------------------------------------------------------

grid <- data.frame(
  W = seq(-2.5, 2.5, length.out = 200)
)

grid$estimated_contrast <- predict.LMTPDRLearner(
  dr_fit,
  newdata = grid,
  type = "contrast"
)

head(grid)
```

# Compare the estimated curve to truth

```{r}
# Plot estimated curve against Monte Carlo truth ---------------------------

ggplot() +
  geom_point(
    data = truth_curve,
    aes(x = W, y = truth),
    alpha = 0.35
  ) +
  geom_line(
    data = grid,
    aes(x = W, y = estimated_contrast),
    linewidth = 1
  ) +
  labs(
    x = "Baseline W",
    y = "Conditional LMTP contrast",
    title = "DR-learner estimate of a smooth conditional LMTP effect",
    subtitle = "Points: large-sample Monte Carlo truth; line: estimated DR-learner curve"
  )
```

# Optional: compare to a simple linear meta-learner

A linear meta-learner is misspecified for this DGP. This comparison is meant to show why the meta-learner matters.

```{r}
# Linear meta-learner comparison ------------------------------------------

dr_fit_lm <- run_drlearner_for_subgroup_LMTP(
  ds = ds,
  nuisance_factory = nf_shift,
  fluctuation = fluctuation,
  metalearner = make_lm_metalearner(),
  conditioning_cols = "W",
  A_type = "continuous"
)

grid$linear_meta_contrast <- predict.LMTPDRLearner(
  dr_fit_lm,
  newdata = grid,
  type = "contrast"
)

ggplot() +
  geom_point(
    data = truth_curve,
    aes(x = W, y = truth),
    alpha = 0.35
  ) +
  geom_line(
    data = grid,
    aes(x = W, y = estimated_contrast),
    linewidth = 1
  ) +
  geom_line(
    data = grid,
    aes(x = W, y = linear_meta_contrast),
    linewidth = 1,
    linetype = "dashed"
  ) +
  labs(
    x = "Baseline W",
    y = "Conditional LMTP contrast",
    title = "Flexible versus linear meta-learning",
    subtitle = "Solid: flexible meta-learner; dashed: linear meta-learner; points: Monte Carlo truth"
  )
```



# Other Approaches

```{r}
# Smooth meta-learners -----------------------------------------------------

# The DR-learner estimates the conditional contrast by regressing the
# DR pseudo-outcome contrast on W. Since this DGP was designed to have a
# smooth nonlinear relationship in W, we compare several smooth meta-learners:
#
#   1. linear lm:       contrast ~ W
#   2. quadratic lm:    contrast ~ W + W^2
#   3. spline lm:       contrast ~ ns(W, df = 5)
#   4. loess:           local polynomial smoothing
#
# The quadratic lm should do especially well here because the true curve has
# a quadratic component.

library(splines)

# Linear meta-learner.
metalearner_linear <- make_lm_metalearner(
  formula = `..dr_pseudooutcome` ~ W
)

# Quadratic meta-learner.
metalearner_quadratic <- make_lm_metalearner(
  formula = `..dr_pseudooutcome` ~ W + I(W^2)
)

# Natural spline meta-learner.
metalearner_spline <- make_lm_metalearner(
  formula = `..dr_pseudooutcome` ~ splines::ns(W, df = 5)
)
```

```{r}
# Loess meta-learner -------------------------------------------------------

make_loess_metalearner <- function(formula = NULL,
                                   span = 0.75,
                                   degree = 2,
                                   ...) {
  loess_args <- list(...)

  function(data, outcome_col, covariate_cols) {
    if (length(covariate_cols) != 1L) {
      stop("`make_loess_metalearner()` currently expects exactly one conditioning covariate.")
    }

    if (is.null(formula)) {
      formula <- stats::as.formula(
        paste(outcome_col, "~", covariate_cols[[1]])
      )
    }

    fit <- do.call(
      stats::loess,
      c(
        list(
          formula = formula,
          data = data,
          span = span,
          degree = degree,
          control = stats::loess.control(surface = "direct")
        ),
        loess_args
      )
    )

    list(
      fit = fit,
      predict = function(newdata) {
        as.numeric(stats::predict(fit, newdata = newdata))
      }
    )
  }
}

metalearner_loess <- make_loess_metalearner(
  formula = `..dr_pseudooutcome` ~ W,
  span = 0.65,
  degree = 2
)
```


```{r}
# Fit DR-learners with smooth meta-learners --------------------------------

dr_fit_linear <- run_drlearner_for_subgroup_LMTP(
  ds = ds,
  nuisance_factory = nf_shift,
  fluctuation = fluctuation,
  metalearner = metalearner_linear,
  conditioning_cols = "W",
  A_type = "continuous"
)

# Recreate the nuisance factory before each fit if your factory stores fitted
# vectors internally. This avoids accidentally reusing mutated state.
nf_shift_quad <- LMTPNuisanceFactory$new(
  policy_seq = policy_shift,
  m_learners = m_learners,
  g_learners = g_learners,
  truncate_density = 1e-12,
  truncate_ratio = c(1e-6, 1e6)
)

dr_fit_quadratic <- run_drlearner_for_subgroup_LMTP(
  ds = ds,
  nuisance_factory = nf_shift_quad,
  fluctuation = fluctuation,
  metalearner = metalearner_quadratic,
  conditioning_cols = "W",
  A_type = "continuous"
)

nf_shift_spline <- LMTPNuisanceFactory$new(
  policy_seq = policy_shift,
  m_learners = m_learners,
  g_learners = g_learners,
  truncate_density = 1e-12,
  truncate_ratio = c(1e-6, 1e6)
)

dr_fit_spline <- run_drlearner_for_subgroup_LMTP(
  ds = ds,
  nuisance_factory = nf_shift_spline,
  fluctuation = fluctuation,
  metalearner = metalearner_spline,
  conditioning_cols = "W",
  A_type = "continuous"
)

nf_shift_loess <- LMTPNuisanceFactory$new(
  policy_seq = policy_shift,
  m_learners = m_learners,
  g_learners = g_learners,
  truncate_density = 1e-12,
  truncate_ratio = c(1e-6, 1e6)
)

dr_fit_loess <- run_drlearner_for_subgroup_LMTP(
  ds = ds,
  nuisance_factory = nf_shift_loess,
  fluctuation = fluctuation,
  metalearner = metalearner_loess,
  conditioning_cols = "W",
  A_type = "continuous"
)
```

```{r}
# Predict conditional contrast curves --------------------------------------

grid <- data.frame(
  W = seq(-2.5, 2.5, length.out = 200)
)

grid$linear <- predict.LMTPDRLearner(
  dr_fit_linear,
  newdata = grid,
  type = "contrast"
)

grid$quadratic <- predict.LMTPDRLearner(
  dr_fit_quadratic,
  newdata = grid,
  type = "contrast"
)

grid$spline <- predict.LMTPDRLearner(
  dr_fit_spline,
  newdata = grid,
  type = "contrast"
)

grid$loess <- predict.LMTPDRLearner(
  dr_fit_loess,
  newdata = grid,
  type = "contrast"
)

plot_df <- rbind(
  data.frame(W = grid$W, estimate = grid$linear, method = "Linear lm"),
  data.frame(W = grid$W, estimate = grid$quadratic, method = "Quadratic lm"),
  data.frame(W = grid$W, estimate = grid$spline, method = "Natural spline"),
  data.frame(W = grid$W, estimate = grid$loess, method = "Loess")
)

ggplot() +
  geom_point(
    data = truth_curve,
    aes(x = W, y = truth),
    alpha = 0.80
  ) +
  geom_line(
    data = plot_df,
    aes(x = W, y = estimate, linetype = method, color = method),
    linewidth = 2
  ) +
  geom_rug(
    data = df,
    aes(x = W),
    alpha = .3) + 
  see::scale_color_okabeito() + 
  theme_minimal() + 
  labs(
    x = "Baseline W",
    y = "Conditional LMTP contrast",
    title = "DR-learner estimates with smooth meta-learners",
    subtitle = "Points: Monte Carlo truth; lines: second-stage DR-learner fits"
  )
```


```{r}
# Compare methods by RMSE against binned truth -----------------------------

truth_eval <- truth_curve
pred_eval <- data.frame(
  W = truth_eval$W
)

pred_eval$linear <- predict(
  dr_fit_linear,
  newdata = pred_eval,
  type = "contrast"
)

pred_eval$quadratic <- predict(
  dr_fit_quadratic,
  newdata = pred_eval,
  type = "contrast"
)

pred_eval$spline <- predict(
  dr_fit_spline,
  newdata = pred_eval,
  type = "contrast"
)

pred_eval$loess <- predict(
  dr_fit_loess,
  newdata = pred_eval,
  type = "contrast"
)

rmse <- data.frame(
  method = c("Linear lm", "Quadratic lm", "Natural spline", "Loess"),
  rmse = c(
    sqrt(mean((pred_eval$linear - truth_eval$truth)^2)),
    sqrt(mean((pred_eval$quadratic - truth_eval$truth)^2)),
    sqrt(mean((pred_eval$spline - truth_eval$truth)^2)),
    sqrt(mean((pred_eval$loess - truth_eval$truth)^2))
  )
)

rmse[order(rmse$rmse), ]
```



