---
title: "Intro to Modified Treatment Policies"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

The `{modulatR}` package represents a longitudinal modified treatment policy
(LMTP) as a sequence of single-time modified treatment policies (MTPs), one
for each treatment time point. The key idea is that a single-time policy acts
on the treatment at one time point given the history available at that time.

An MTP object is therefore a single-time transformation object. It contains
the map from an observed treatment value $A_t$ to a shifted value $A_t^d$,
and, how to compute the induced post-policy density
$g_t^d(a_t | h_t)$ from the observed conditional density $g_t(a_t | h_t)$
using the transformation of variables formula.

A `LMTPPolicySequence` simply stores one such object per time point.

This vignette focuses on just the single timepoint policy objects themselves
rather than their longitudinal versions in `LMTPPolicySequence`.

### A simple continuous modified treatment policy

```{r setup}
library(modulatR)
```

We start with a single-time continuous treatment A having a lognormal
distribution. Consider the policy

$$
d(A, L) = \begin{cases}
a - 5, & a \geq 5, \\
a, & a < 5.
\end{cases}
$$

This is a useful example because:

  - it is easy to visualize;
  - it is piecewise invertible;
  - we will use it to show how an MTP can alter a distribution without being a
static intervention.

Below we simulate from a lognormal treatment distribution and then define this
policy as a two-region MTP.

```{r}
library(ggplot2)

set.seed(2026)

n <- 5000
L <- data.frame(L1 = rnorm(n))
A <- rlnorm(n, meanlog = 1.6, sdlog = 0.45)

mtp_shift_down_5 <- MTP$new(
  treatment_type = "continuous",
  region_predicates = list(
    function(A, H) A >= 5,
    function(A, H) A < 5
  ),
  policy_pieces = list(
    function(A, H) A - 5,
    function(A, H) A
  ),
  inverse_map_pieces = list(
    function(A_star, H) A_star + 5,
    function(A_star, H) A_star
  ),
  inverse_deriv_pieces = list(
    function(A_star, H) rep(1, length(A_star)),
    function(A_star, H) rep(1, length(A_star))
  ),
  name = "shift_down_5_if_A_ge_5"
)

mtp_shift_down_5

A_star <- mtp_shift_down_5$apply_policy(A, L)
```

Applying the policy directly:

```{r}
head(data.frame(
  A = round(A[1:6], 2),
  A_star = round(A_star[1:6], 2)
))
```

### Visualizing the transformation 


```{r}
dist_df <- rbind(
  data.frame(value = A, which = "Observed A"),
  data.frame(value = A_star, which = "Shifted A*")
)

ggplot(dist_df, aes(x = value)) +
  geom_histogram(bins = 60) +
  facet_wrap(~ which, ncol = 1, scales = "free_y") +
  labs(
    title = "Observed and post-policy treatment distributions",
    x = "Treatment value",
    y = "Count"
  ) +
  theme_minimal()
```


```{r}
g_density <- function(A_vec, H_df) {
  dlnorm(A_vec, meanlog = 1.6, sdlog = 1.1)
}

grid <- seq(0.001, 30, length.out = 500)

gd_grid <- mtp_shift_down_5$gd_from_density(
  A_vec = grid,
  H_df = data.frame(L1 = rep(0, length(grid))),
  density_fun = g_density
)

density_df <- rbind(
  data.frame(x = grid, density = g_density(grid, data.frame(L1 = 0)), which = "Observed g"),
  data.frame(x = grid, density = gd_grid, which = "Post-policy g^d")
)

ggplot(density_df, aes(x = x, y = density, linetype = which, group = which)) +
  geom_line(linewidth = 1) +
  labs(
    title = "Observed density and post-policy density",
    x = "Treatment value",
    y = "Density",
    linetype = "Distribution"
  ) +
  scale_linetype_manual(values = c('solid', 'dotdash')) + 
  theme_minimal() + 
  xlim(c(0, 10))
```
