---
title: "Estimating a 12-timepoint LMTP under a known data-generating process"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimating a 12-timepoint LMTP under a known data-generating process}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)

set.seed(2026)
library(modulatR)
library(ggplot2)
```

# Overview

This vignette estimates a longitudinal modified treatment policy (LMTP) with
12 treatment times. The observed data have the structure

\[
  O =
  (W, L_1, A_1, L_2, A_2, \ldots, L_{12}, A_{12}, Y),
\]

where \(W\) is a baseline covariate, \(L_t\) is a time-varying covariate,
\(A_t\) is a continuous treatment, and \(Y\) is a final continuous outcome.

We consider the additive modified treatment policy

\[
  d_t(a_t, h_t) = a_t + 1,
  \qquad t = 1, \ldots, 12.
\]

The target parameter is

\[
  \mathbb{E}(Y^{\bar d}),
\]

and we compare it to the observed-data mean \(\mathbb{E}(Y)\). Since the
data-generating process is known, we also compute the truth by Monte Carlo.

# Data-generating process

The DGP is intentionally linear and correctly specified by the GLM nuisance
models used below. This makes the vignette a clean diagnostic for whether the
12-timepoint LMTP machinery is behaving as expected.

At each time \(t\),

\[
  L_t =
  0.35 W + 0.25 L_{t-1} + 0.50 A_{t-1} + \varepsilon_{L,t},
\]

and

\[
  A_t =
  0.30 W + 0.45 L_t + 0.25 A_{t-1} + \varepsilon_{A,t},
\]

with \(L_0 = 0\) and \(A_0 = 0\). The outcome is

\[
  Y =
  1 + 0.5 W
    + \sum_{t = 1}^{12} \beta_{L,t} L_t
    + \sum_{t = 1}^{12} \beta_{A,t} A_t
    + \varepsilon_Y.
\]

The treatment effect coefficients decay mildly over time, so that earlier
treatments have somewhat larger effects than later treatments.

```{r}
library(nadir)

# -------------------------------------------------------------------------
# Data generator
# -------------------------------------------------------------------------

generate_lmtp_data <- function(n,
                               tau = 12,
                               delta = NULL,
                               seed = NULL) {
  if (!is.null(seed)) set.seed(seed)

  W <- rnorm(n)

  eL <- matrix(rnorm(n * tau), nrow = n, ncol = tau)
  eA <- matrix(rnorm(n * tau), nrow = n, ncol = tau)
  eY <- rnorm(n)

  L <- matrix(NA_real_, nrow = n, ncol = tau)
  A <- matrix(NA_real_, nrow = n, ncol = tau)

  colnames(L) <- paste0("L", seq_len(tau))
  colnames(A) <- paste0("A", seq_len(tau))

  prev_L <- rep(0, n)
  prev_A <- rep(0, n)

  for (t in seq_len(tau)) {
    L[, t] <- 0.35 * W + .65 * prev_L + .45 * prev_A + eL[, t]
    A[, t] <- 0.30 * W + .91 * L[, t] + .95 * prev_A + eA[, t]

    if (!is.null(delta)) {
      A[, t] <- A[, t] + delta
    }

    prev_L <- L[, t]
    prev_A <- A[, t]
  }

  beta_L <- 0.20 + 0.10 * exp(-0.15 * (seq_len(tau) - 1))
  beta_A <- 0.60 + 0.60 * exp(-0.15 * (seq_len(tau) - 1))

  Y <- 1 +
    0.50 * W +
    as.numeric(L %*% beta_L) +
    as.numeric(A %*% beta_A) +
    eY

  data.frame(
    W = W,
    as.data.frame(L),
    as.data.frame(A),
    Y = Y,
    check.names = FALSE
  )
}
```

# Truth by Monte Carlo

Because this is a known DGP, we can approximate the truth by simulating a large
sample under the observed law and under the shifted law. The shifted law uses
the same structural equations but applies \(A_t \mapsto A_t + 1\) at every
timepoint.

For package vignettes, we use a moderately large Monte Carlo sample to keep
runtime manageable. If you are running this interactively, increasing
`n_truth` will reduce Monte Carlo error.

```{r}
# -------------------------------------------------------------------------
# Truth by large Monte Carlo sample
# -------------------------------------------------------------------------

set.seed(1)

tau <- 12
delta <- 1
n_truth <- 2e5

truth_obs <- generate_lmtp_data(
  n = n_truth,
  tau = tau,
  delta = NULL
)

truth_shift <- generate_lmtp_data(
  n = n_truth,
  tau = tau,
  delta = delta
)

truth <- c(
  EY = mean(truth_obs$Y),
  EYd = mean(truth_shift$Y),
  diff = mean(truth_shift$Y - truth_obs$Y)
)

truth
```

# Observed data

Now we generate one observed dataset and wrap it in an `LMTPData` object. The
important bookkeeping step is that `A_cols` is a character vector of treatment
names, while `L_cols` is a list whose \(t\)-th element gives the covariates
available at time \(t\).

```{r}
# -------------------------------------------------------------------------
# Observed data
# -------------------------------------------------------------------------

set.seed(2)

n <- 8000

df <- generate_lmtp_data(
  n = n,
  tau = tau,
  delta = NULL
)

A_cols <- paste0("A", seq_len(tau))
L_cols <- as.list(paste0("L", seq_len(tau)))

ds <- LMTPData$new(
  data = df,
  A_cols = A_cols,
  L_cols = L_cols,
  W_cols = "W",
  Y_col = "Y"
)

ds
```

We can inspect the histories at early and late timepoints. The history
\(H_t\) contains baseline covariates, time-varying covariates through time
\(t\), and prior treatments through time \(t - 1\).

```{r}
head(ds$H(1))
head(ds$H(12))
head(ds$AH(12))
```

# Policies

We define two policy sequences. The first adds one unit to every treatment at
every time. The second is the identity policy, which estimates the observed-law
mean using the same machinery.

```{r}
# -------------------------------------------------------------------------
# Policies
# -------------------------------------------------------------------------

policy_shift <- repeat_policy_over_time(
  mtp_additive_shift(delta = delta),
  tau = ds$tau()
)

policy_identity <- identity_policy(
  A_type = "continuous",
  tau = ds$tau()
)

policy_shift
policy_identity
```

It is often useful to check the implied treatment shifts directly.

```{r}
shifted_treatments <- policy_shift$apply_to_data(ds)

head(shifted_treatments[, c(
  ds$id_col,
  "A1", "A1_star",
  "A6", "A6_star",
  "A12", "A12_star"
)])
```

# Learners

For this simple known DGP, correctly specified GLMs are adequate. We use GLMs
for the sequential outcome regressions and `nadir::lnr_glm_density` for the
conditional treatment densities.

```{r}
# -------------------------------------------------------------------------
# Learners
# -------------------------------------------------------------------------

m_learners <- make_glm_m_learners(
  tau = ds$tau(),
  family = stats::gaussian()
)

g_learners <- make_nadir_density_g_learners(
  tau = ds$tau(),
  learner = nadir::lnr_glm_density
)
```

# Nuisance factories and fluctuation model

The nuisance factory stores the policy sequence and the per-time learners used
for the sequential regressions and density ratios. Since the outcome is
continuous, we use a Gaussian fluctuation submodel.

```{r}
# -------------------------------------------------------------------------
# Nuisance factories
# -------------------------------------------------------------------------

nf_shift <- LMTPNuisanceFactory$new(
  policy_seq = policy_shift,
  m_learners = m_learners,
  g_learners = g_learners,
  truncate_density = 1e-12,
  truncate_ratio = c(1e-6, Inf)
)

nf_identity <- LMTPNuisanceFactory$new(
  policy_seq = policy_identity,
  m_learners = m_learners,
  g_learners = g_learners,
  truncate_density = 1e-12,
  truncate_ratio = c(1e-6, Inf)
)

fluctuation <- LMTPFluctuationSubmodel$new(
  family = stats::gaussian(),
  use_intercept = TRUE
)

fluctuation
```

# Fit the TMLEs

We now estimate both \(\mathbb{E}(Y^{\bar d})\) and \(\mathbb{E}(Y)\).

```{r}
# -------------------------------------------------------------------------
# Fit TMLEs
# -------------------------------------------------------------------------

fit_shift <- run_tmle_for_LMTP(
  ds = ds,
  nuisance_factory = nf_shift,
  fluctuation = fluctuation
)

fit_identity <- run_tmle_for_LMTP(
  ds = ds,
  nuisance_factory = nf_identity,
  fluctuation = fluctuation
)

fit_shift
fit_identity
```

# Difference contrast

Finally, we contrast the shifted-policy mean against the identity-policy mean.

```{r}
# -------------------------------------------------------------------------
# Difference contrast
# -------------------------------------------------------------------------

fit_diff <- lmtp_delta_contrast(
  fit1 = fit_shift,
  fit0 = fit_identity,
  transform = "difference"
)

comparison <- data.frame(
  parameter = c("E[Y]", "E[Y^d]", "E[Y^d] - E[Y]"),
  truth = c(
    truth["EY"],
    truth["EYd"],
    truth["diff"]
  ),
  estimate = c(
    fit_identity$estimate,
    fit_shift$estimate,
    fit_diff$estimate
  ),
  se = c(
    fit_identity$se,
    fit_shift$se,
    fit_diff$se
  )
)

comparison$error <- comparison$estimate - comparison$truth

comparison
fit_diff
```

## Monte Carlo distribution of the estimators

To visualize the sampling distribution of the estimators, we now repeat the
entire estimation procedure over many independently generated datasets. This
lets us plot the empirical distribution of the estimated parameters and compare
it to the Monte Carlo truth.

For vignette speed, we keep the number of repetitions moderate. If you run this
interactively, you may wish to increase `B`.

```{r}
fit_lmtp_once <- function(n,
                          tau = 12,
                          delta = 1) {

  df <- generate_lmtp_data(
    n = n,
    tau = tau,
    delta = NULL
  )

  A_cols <- paste0("A", seq_len(tau))
  L_cols <- as.list(paste0("L", seq_len(tau)))

  ds <- LMTPData$new(
    data = df,
    A_cols = A_cols,
    L_cols = L_cols,
    W_cols = "W",
    Y_col = "Y"
  )

  policy_shift <- repeat_policy_over_time(
    mtp_additive_shift(delta = delta),
    tau = ds$tau()
  )

  policy_identity <- identity_policy(
    A_type = "continuous",
    tau = ds$tau()
  )

  m_learners <- make_glm_m_learners(
    tau = ds$tau(),
    family = stats::gaussian()
  )

  g_learners <- make_nadir_density_g_learners(
    tau = ds$tau(),
    learner = nadir::lnr_glm_density
  )

  nf_shift <- LMTPNuisanceFactory$new(
    policy_seq = policy_shift,
    m_learners = m_learners,
    g_learners = g_learners,
    truncate_density = 1e-12,
    truncate_ratio = c(1e-6, Inf)
  )

  nf_identity <- LMTPNuisanceFactory$new(
    policy_seq = policy_identity,
    m_learners = m_learners,
    g_learners = g_learners,
    truncate_density = 1e-12,
    truncate_ratio = c(1e-6, Inf)
  )

  fluctuation <- LMTPFluctuationSubmodel$new(
    family = stats::gaussian(),
    use_intercept = TRUE
  )

  fit_shift <- run_tmle_for_LMTP(
    ds = ds,
    nuisance_factory = nf_shift,
    fluctuation = fluctuation
  )

  fit_identity <- run_tmle_for_LMTP(
    ds = ds,
    nuisance_factory = nf_identity,
    fluctuation = fluctuation
  )

  fit_diff <- lmtp_delta_contrast(
    fit1 = fit_shift,
    fit0 = fit_identity,
    transform = "difference"
  )

  c(
    "E[Y]" = fit_identity$estimate,
    "E[Y^d]" = fit_shift$estimate,
    "E[Y^d] - E[Y]" = fit_diff$estimate
  )
}
```

```{r}
set.seed(1234)

B <- 50

mc_estimates <- replicate(
  B,
  fit_lmtp_once(
    n = n,
    tau = tau,
    delta = delta
  )
)

mc_df <- data.frame(
  replicate = rep(seq_len(B), each = nrow(mc_estimates)),
  parameter = factor(
    rep(rownames(mc_estimates), times = ncol(mc_estimates)),
    levels = rownames(mc_estimates)
  ),
  estimate = as.vector(mc_estimates)
)

truth_df <- data.frame(
  parameter = factor(
    c("E[Y]", "E[Y^d]", "E[Y^d] - E[Y]"),
    levels = levels(mc_df$parameter)
  ),
  truth = c(
    truth["EY"],
    truth["EYd"],
    truth["diff"]
  )
)

head(mc_df)
truth_df
```

```{r}
ggplot(mc_df, aes(x = parameter, y = estimate)) +
  geom_jitter(
    width = 0.12,
    height = 0,
    alpha = 0.35,
    size = 1.6,
    color = "gray35"
  ) +
  geom_boxplot(
    width = 0.45,
    alpha = 0.35,
    outlier.shape = NA,
    fill = "skyblue3",
    color = "gray15"
  ) +
  geom_point(
    data = truth_df,
    aes(x = parameter, y = truth),
    inherit.aes = FALSE,
    shape = 17,
    size = 4.8,
    color = "#D55E00"
  ) +
  labs(
    x = NULL,
    y = "Estimate",
    title = "Sampling distribution of LMTP estimates across simulated datasets",
    subtitle = "Triangles indicate Monte Carlo truth"
  ) +
  theme_minimal(base_size = 12)
```

# Remarks

This example is deliberately favorable to the estimator. The nuisance models
are correctly specified, the policy is a simple additive shift, and the
treatment densities are smooth and continuous. In more realistic applications,
one would usually replace the GLM learners with richer learners and use
cross-fitting.

For example, one could wrap the same per-time learners with
`crossfit_lmtp_learner()` before passing them to the nuisance factory. This is
often preferable when using flexible machine-learning nuisance estimators.

```{r, eval = FALSE}
m_learners_cf <- lapply(
  m_learners,
  crossfit_lmtp_learner,
  V = 5
)

g_learners_cf <- lapply(
  g_learners,
  crossfit_lmtp_learner,
  V = 5
)

nf_shift_cf <- LMTPNuisanceFactory$new(
  policy_seq = policy_shift,
  m_learners = m_learners_cf,
  g_learners = g_learners_cf,
  truncate_density = 1e-12,
  truncate_ratio = c(1e-6, Inf)
)

fit_shift_cf <- run_tmle_for_LMTP(
  ds = ds,
  nuisance_factory = nf_shift_cf,
  fluctuation = fluctuation
)
```
