---
title: "Estimating longitudinal ATEs for a binary treatment"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimating longitudinal ATEs for a binary treatment}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)

set.seed(2026)

library(modulatR)
library(ggplot2)
```

# Overview

This vignette shows how to use `{modulatR}` to estimate a longitudinal average
treatment effect for a binary treatment.

The observed data have the longitudinal structure

\[
  O =
  (W, L_1, A_1, L_2, A_2, \ldots, L_\tau, A_\tau, Y),
\]

where \(W\) is a baseline covariate, \(L_t\) is a time-varying covariate,
\(A_t \in \{0,1\}\) is a binary treatment, and \(Y\) is a final continuous
outcome.

The estimand is the static-regime contrast

\[
  \psi =
  \mathbb{E}\{Y^{\bar A = 1}\}
  -
  \mathbb{E}\{Y^{\bar A = 0}\}.
\]

Although `{modulatR}` is organized around modified treatment policies, a static
binary treatment regime is a valid discrete policy. For example,

\[
  d_t(A_t, H_t) = 1
\]

is the policy that sets treatment to one at time \(t\), regardless of the
observed treatment or history.

# Data-generating process

We simulate a longitudinal dataset with binary treatments. Treatment at each
time depends on the current covariate history, and later covariates depend on
earlier treatment.

```{r}
# -------------------------------------------------------------------------
# Data generator
# -------------------------------------------------------------------------

generate_binary_lmtp_data <- function(n,
                                      tau = 6,
                                      static_a = NULL,
                                      seed = NULL) {
  if (!is.null(seed)) set.seed(seed)

  W <- rnorm(n)

  L <- matrix(NA_real_, nrow = n, ncol = tau)
  A <- matrix(NA_integer_, nrow = n, ncol = tau)

  colnames(L) <- paste0("L", seq_len(tau))
  colnames(A) <- paste0("A", seq_len(tau))

  prev_L <- rep(0, n)
  prev_A <- rep(0, n)

  for (t in seq_len(tau)) {
    L[, t] <-
      0.35 * W +
      0.40 * prev_L +
      0.70 * prev_A +
      rnorm(n)

    p_A_t <- plogis(
      -0.15 +
        0.35 * W +
        0.65 * L[, t] -
        0.25 * prev_A
    )

    if (is.null(static_a)) {
      A[, t] <- rbinom(n, size = 1, prob = p_A_t)
    } else {
      A[, t] <- static_a
    }

    prev_L <- L[, t]
    prev_A <- A[, t]
  }

  beta_L <- 0.15 + 0.05 * exp(-0.20 * (seq_len(tau) - 1))
  beta_A <- 0.50 + 0.35 * exp(-0.20 * (seq_len(tau) - 1))

  Y <-
    1 +
    0.40 * W +
    as.numeric(L %*% beta_L) +
    as.numeric(A %*% beta_A) +
    rnorm(n)

  data.frame(
    W = W,
    as.data.frame(L),
    as.data.frame(A),
    Y = Y,
    check.names = FALSE
  )
}
```

# Monte Carlo truth

Because the data-generating process is known, we approximate the truth using a
large Monte Carlo sample. We simulate under the always-treated regime and under
the never-treated regime.

```{r}
# -------------------------------------------------------------------------
# Truth by large Monte Carlo sample
# -------------------------------------------------------------------------

set.seed(1)

tau <- 6
n_truth <- 2e5

truth_one <- generate_binary_lmtp_data(
  n = n_truth,
  tau = tau,
  static_a = 1
)

truth_zero <- generate_binary_lmtp_data(
  n = n_truth,
  tau = tau,
  static_a = 0
)

truth <- c(
  EY1 = mean(truth_one$Y),
  EY0 = mean(truth_zero$Y),
  ATE = mean(truth_one$Y) - mean(truth_zero$Y)
)

truth
```

# Observed data

We now generate one observed dataset under the natural treatment process.

```{r}
# -------------------------------------------------------------------------
# Observed data
# -------------------------------------------------------------------------

set.seed(2)

n <- 8000

df <- generate_binary_lmtp_data(
  n = n,
  tau = tau,
  static_a = NULL
)

A_cols <- paste0("A", seq_len(tau))
L_cols <- as.list(paste0("L", seq_len(tau)))

ds <- LMTPData$new(
  data = df,
  A_cols = A_cols,
  L_cols = L_cols,
  W_cols = "W",
  Y_col = "Y"
)

ds
```

It is useful to check the treatment prevalence by time.

```{r}
treatment_prevalence <- data.frame(
  time = seq_len(tau),
  treatment = A_cols,
  prevalence = vapply(df[A_cols], mean, numeric(1))
)

treatment_prevalence
```

# Static binary treatment policies

The always-treated policy maps every observed treatment value to one. The
never-treated policy maps every observed treatment value to zero.

```{r}
# -------------------------------------------------------------------------
# Static binary policies
# -------------------------------------------------------------------------

mtp_always_treat <- mtp_discrete(
  map_fun = function(A, H_df) rep(1, length(A)),
  support = c(0, 1),
  name = "always_treat"
)

mtp_never_treat <- mtp_discrete(
  map_fun = function(A, H_df) rep(0, length(A)),
  support = c(0, 1),
  name = "never_treat"
)

policy_always_treat <- repeat_policy_over_time(
  mtp_always_treat,
  tau = ds$tau(),
  name = "always_treat_sequence"
)

policy_never_treat <- repeat_policy_over_time(
  mtp_never_treat,
  tau = ds$tau(),
  name = "never_treat_sequence"
)

policy_always_treat
policy_never_treat
```

We can inspect the policies on the observed data.

```{r}
policy_check_one <- policy_always_treat$apply_to_data(ds)
policy_check_zero <- policy_never_treat$apply_to_data(ds)

head(policy_check_one[, c(ds$id_col, "A1", "A1_star", "A6", "A6_star")])
head(policy_check_zero[, c(ds$id_col, "A1", "A1_star", "A6", "A6_star")])
```

# Learners

For the outcome regressions, we use correctly specified Gaussian GLMs. For the
treatment mechanism ratios, we use the ratio-classification approach through
`lambda_learners`.

The `lambda_learners` estimate a classification nuisance in an augmented dataset
that distinguishes observed rows from intervention-shifted rows. The nuisance
factory then converts this classification regression into density or mass
ratios.

```{r}
# -------------------------------------------------------------------------
# Learners
# -------------------------------------------------------------------------

m_learners <- make_glm_m_learners(
  tau = ds$tau(),
  family = stats::gaussian()
)

lambda_learners <- make_glm_lambda_learners(
  tau = ds$tau(),
  family = stats::binomial()
)
```

# Nuisance factories

We need one nuisance factory for each policy because the policy sequence is part
of the nuisance factory.

```{r}
# -------------------------------------------------------------------------
# Nuisance factories
# -------------------------------------------------------------------------

nf_always_treat <- LMTPNuisanceFactory$new(
  policy_seq = policy_always_treat,
  m_learners = m_learners,
  lambda_learners = lambda_learners,
  clip_lambda_probability = 1e-6,
  truncate_ratio = c(1e-6, 1e6)
)

nf_never_treat <- LMTPNuisanceFactory$new(
  policy_seq = policy_never_treat,
  m_learners = m_learners,
  lambda_learners = lambda_learners,
  clip_lambda_probability = 1e-6,
  truncate_ratio = c(1e-6, 1e6)
)

fluctuation <- LMTPFluctuationSubmodel$new(
  family = stats::gaussian(),
  use_intercept = TRUE
)
```

# Fit the longitudinal TMLEs

We estimate

\[
  \mathbb{E}\{Y^{\bar A = 1}\}
\]

and

\[
  \mathbb{E}\{Y^{\bar A = 0}\}.
\]

```{r}
# -------------------------------------------------------------------------
# Fit TMLEs
# -------------------------------------------------------------------------

fit_always_treat <- run_tmle_for_LMTP(
  ds = ds,
  nuisance_factory = nf_always_treat,
  fluctuation = fluctuation
)

fit_never_treat <- run_tmle_for_LMTP(
  ds = ds,
  nuisance_factory = nf_never_treat,
  fluctuation = fluctuation
)

fit_always_treat
fit_never_treat
```

# Longitudinal ATE contrast

The longitudinal average treatment effect is the difference between the
always-treated and never-treated counterfactual means.

```{r}
# -------------------------------------------------------------------------
# Difference contrast
# -------------------------------------------------------------------------

fit_ate <- lmtp_delta_contrast(
  fit1 = fit_always_treat,
  fit0 = fit_never_treat,
  transform = "difference"
)

comparison <- data.frame(
  parameter = c(
    "E[Y^{Abar = 1}]",
    "E[Y^{Abar = 0}]",
    "ATE"
  ),
  truth = c(
    truth["EY1"],
    truth["EY0"],
    truth["ATE"]
  ),
  estimate = c(
    fit_always_treat$estimate,
    fit_never_treat$estimate,
    fit_ate$estimate
  ),
  se = c(
    fit_always_treat$se,
    fit_never_treat$se,
    fit_ate$se
  )
)

comparison$lower <- comparison$estimate - qnorm(0.975) * comparison$se
comparison$upper <- comparison$estimate + qnorm(0.975) * comparison$se
comparison$error <- comparison$estimate - comparison$truth

comparison
fit_ate
```

# Visual check

```{r}
ggplot(comparison, aes(x = parameter, y = estimate)) +
  geom_errorbar(
    aes(ymin = lower, ymax = upper),
    width = 0.12
  ) +
  geom_point(size = 2.5) +
  geom_point(
    aes(y = truth),
    shape = 17,
    size = 4.5,
    color = "#D55E00"
  ) +
  labs(
    x = NULL,
    y = "Value",
    title = "Longitudinal static-regime estimates",
    subtitle = "Triangles indicate Monte Carlo truth"
  ) +
  theme_minimal(base_size = 12)
```

# Diagnostics for the ratio-classification nuisances

The weights can become unstable when treatment positivity is weak. For a simple
diagnostic, we inspect the cumulative ratio process \(\omega_t\) under each
policy.

```{r}
# -------------------------------------------------------------------------
# Ratio diagnostics
# -------------------------------------------------------------------------

omega_summary <- function(nf, label) {
  data.frame(
    policy = label,
    time = seq_len(nf$tau),
    min = vapply(nf$omega_preds, min, numeric(1)),
    q25 = vapply(nf$omega_preds, quantile, numeric(1), probs = 0.25),
    median = vapply(nf$omega_preds, median, numeric(1)),
    mean = vapply(nf$omega_preds, mean, numeric(1)),
    q75 = vapply(nf$omega_preds, quantile, numeric(1), probs = 0.75),
    max = vapply(nf$omega_preds, max, numeric(1))
  )
}

omega_diagnostics <- rbind(
  omega_summary(nf_always_treat, "always_treat"),
  omega_summary(nf_never_treat, "never_treat")
)

omega_diagnostics
```

```{r}
ggplot(omega_diagnostics, aes(x = time, y = mean, group = policy)) +
  geom_line() +
  geom_point() +
  labs(
    x = "Time",
    y = "Mean cumulative ratio",
    title = "Simple cumulative ratio diagnostic"
  ) +
  theme_minimal(base_size = 12)
```

# Monte Carlo distribution of the estimators

As an optional check, we can repeat the analysis over several independently
generated observed datasets. This gives a rough empirical sampling distribution
for the estimated static-regime means and the ATE.

For a package vignette, keep `B` small. For a simulation study, increase it.

```{r}
fit_binary_lmtp_once <- function(n,
                                 tau = 6) {
  df <- generate_binary_lmtp_data(
    n = n,
    tau = tau,
    static_a = NULL
  )

  A_cols <- paste0("A", seq_len(tau))
  L_cols <- as.list(paste0("L", seq_len(tau)))

  ds <- LMTPData$new(
    data = df,
    A_cols = A_cols,
    L_cols = L_cols,
    W_cols = "W",
    Y_col = "Y"
  )

  mtp_always_treat <- mtp_discrete(
    map_fun = function(A, H_df) rep(1, length(A)),
    support = c(0, 1),
    name = "always_treat"
  )

  mtp_never_treat <- mtp_discrete(
    map_fun = function(A, H_df) rep(0, length(A)),
    support = c(0, 1),
    name = "never_treat"
  )

  policy_always_treat <- repeat_policy_over_time(
    mtp_always_treat,
    tau = ds$tau(),
    name = "always_treat_sequence"
  )

  policy_never_treat <- repeat_policy_over_time(
    mtp_never_treat,
    tau = ds$tau(),
    name = "never_treat_sequence"
  )

  m_learners <- make_glm_m_learners(
    tau = ds$tau(),
    family = stats::gaussian()
  )

  lambda_learners <- make_glm_lambda_learners(
    tau = ds$tau(),
    family = stats::binomial()
  )

  nf_always_treat <- LMTPNuisanceFactory$new(
    policy_seq = policy_always_treat,
    m_learners = m_learners,
    lambda_learners = lambda_learners,
    clip_lambda_probability = 1e-6,
    truncate_ratio = c(1e-6, 1e6)
  )

  nf_never_treat <- LMTPNuisanceFactory$new(
    policy_seq = policy_never_treat,
    m_learners = m_learners,
    lambda_learners = lambda_learners,
    clip_lambda_probability = 1e-6,
    truncate_ratio = c(1e-6, 1e6)
  )

  fluctuation <- LMTPFluctuationSubmodel$new(
    family = stats::gaussian(),
    use_intercept = TRUE
  )

  fit_always_treat <- run_tmle_for_LMTP(
    ds = ds,
    nuisance_factory = nf_always_treat,
    fluctuation = fluctuation
  )

  fit_never_treat <- run_tmle_for_LMTP(
    ds = ds,
    nuisance_factory = nf_never_treat,
    fluctuation = fluctuation
  )

  fit_ate <- lmtp_delta_contrast(
    fit1 = fit_always_treat,
    fit0 = fit_never_treat,
    transform = "difference"
  )

  c(
    "E[Y^{Abar = 1}]" = fit_always_treat$estimate,
    "E[Y^{Abar = 0}]" = fit_never_treat$estimate,
    "ATE" = fit_ate$estimate
  )
}
```

```{r}
set.seed(1234)

B <- 30

mc_estimates <- replicate(
  B,
  fit_binary_lmtp_once(
    n = n,
    tau = tau
  )
)

mc_df <- data.frame(
  replicate = rep(seq_len(B), each = nrow(mc_estimates)),
  parameter = factor(
    rep(rownames(mc_estimates), times = ncol(mc_estimates)),
    levels = rownames(mc_estimates)
  ),
  estimate = as.vector(mc_estimates)
)

truth_df <- data.frame(
  parameter = factor(
    c("E[Y^{Abar = 1}]", "E[Y^{Abar = 0}]", "ATE"),
    levels = levels(mc_df$parameter)
  ),
  truth = c(
    truth["EY1"],
    truth["EY0"],
    truth["ATE"]
  )
)

head(mc_df)
truth_df
```

```{r}
ggplot(mc_df, aes(x = parameter, y = estimate)) +
  geom_jitter(
    width = 0.12,
    height = 0,
    alpha = 0.35,
    size = 1.6,
    color = "gray35"
  ) +
  geom_boxplot(
    width = 0.45,
    alpha = 0.35,
    outlier.shape = NA,
    fill = "skyblue3",
    color = "gray15"
  ) +
  geom_point(
    data = truth_df,
    aes(x = parameter, y = truth),
    inherit.aes = FALSE,
    shape = 17,
    size = 4.8,
    color = "#D55E00"
  ) +
  labs(
    x = NULL,
    y = "Estimate",
    title = "Sampling distribution of longitudinal binary-treatment estimates",
    subtitle = "Triangles indicate Monte Carlo truth"
  ) +
  theme_minimal(base_size = 12)
```
