---
title: "Using ratio classification for LMTP density ratios"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using ratio classification for LMTP density ratios}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)

set.seed(2026)

library(modulatR)
library(nadir)

```

For LMTP estimation, the density ratio is
$$r_t(a_t,h_t)
=
\frac{g_t^d(a_t \mid h_t)}
     {g_t(a_t \mid h_t)}.$$
     
Rather than estimating both densities directly, the ratio-classification trick builds an augmented dataset containing observed rows and policy-shifted rows. A binary learner estimates
$$\Pr(\Lambda = 1 \mid A_t,H_t),$$
where
$\Lambda =0$ denotes observed rows and
$\Lambda = 1$ denotes policy-shifted rows. The nuisance factory then converts this probability to a density-ratio estimate by

$$\hat r_t(a_t,h_t)
=
\frac{\hat \Pr(\Lambda=1 \mid a_t,h_t)}
     {1-\hat \Pr(\Lambda=1 \mid a_t,h_t)}.$$
     
This is the classification-based density-ratio strategy described by Díaz et al. for LMTP estimation.

```{r}
# Data-generating process --------------------------------------------------

set.seed(1)

n <- 1500

W  <- rnorm(n)
L1 <- 0.5 * W + rnorm(n)
A1 <- 0.6 * W + 0.8 * L1 + rnorm(n)
L2 <- 0.4 * W + 0.7 * A1 + 0.6 * L1 + rnorm(n)
A2 <- 0.3 * W + 0.5 * L2 + rnorm(n)

Y <- 1 + 0.5 * W + 0.4 * L1 + 0.3 * L2 + 1.2 * A1 + 0.8 * A2 + rnorm(n)

df <- data.frame(W, L1, A1, L2, A2, Y)

ds <- LMTPData$new(
  data = df,
  A_cols = c("A1", "A2"),
  L_cols = list("L1", "L2"),
  W_cols = "W",
  Y_col = "Y"
)

ds
```

# Define an Additive Policy 
```{r}

policy_shift <- repeat_policy_over_time(
  mtp_additive_shift(delta = 1),
  tau = ds$tau()
)

policy_shift
```

# Learners

```{r}
m_learners <- make_glm_m_learners(
  tau = ds$tau(),
  family = stats::gaussian()
)

lambda_learners <- make_glm_lambda_learners(
  tau = ds$tau(),
  family = stats::binomial()
)
```

# Check the ratio-classification machinery directly

This block trains only the ratio nuisance functions and inspects the resulting
$\hat{\lambda}_t$, $\hat r_t$ and cumulative ratio  $\hat \omega_t = \prod_{s\leq t} \hat r_s$. 

```{r}
nf_lambda_check <- LMTPNuisanceFactory$new(
  policy_seq = policy_shift,
  m_learners = m_learners,
  lambda_learners = lambda_learners,
  clip_lambda_probability = 1e-6,
  truncate_ratio = c(1e-6, 1e6)
)

nf_lambda_check$train_ratios(ds)

ratio_diagnostics <- data.frame(
  time = rep(seq_len(ds$tau()), each = 3),
  quantity = rep(c("lambda_obs", "r_obs", "omega"), times = ds$tau()),
  mean = c(
    mean(nf_lambda_check$lambda_obs_preds[[1]]),
    mean(nf_lambda_check$r_obs_preds[[1]]),
    mean(nf_lambda_check$omega_preds[[1]]),
    mean(nf_lambda_check$lambda_obs_preds[[2]]),
    mean(nf_lambda_check$r_obs_preds[[2]]),
    mean(nf_lambda_check$omega_preds[[2]])
  ),
  min = c(
    min(nf_lambda_check$lambda_obs_preds[[1]]),
    min(nf_lambda_check$r_obs_preds[[1]]),
    min(nf_lambda_check$omega_preds[[1]]),
    min(nf_lambda_check$lambda_obs_preds[[2]]),
    min(nf_lambda_check$r_obs_preds[[2]]),
    min(nf_lambda_check$omega_preds[[2]])
  ),
  max = c(
    max(nf_lambda_check$lambda_obs_preds[[1]]),
    max(nf_lambda_check$r_obs_preds[[1]]),
    max(nf_lambda_check$omega_preds[[1]]),
    max(nf_lambda_check$lambda_obs_preds[[2]]),
    max(nf_lambda_check$r_obs_preds[[2]]),
    max(nf_lambda_check$omega_preds[[2]])
  )
)

ratio_diagnostics
```

# shape checks / diagnostics

```{r}
stopifnot(length(nf_lambda_check$lambda_obs_preds) == ds$tau())
stopifnot(length(nf_lambda_check$r_obs_preds) == ds$tau())
stopifnot(length(nf_lambda_check$r_d_preds) == ds$tau())
stopifnot(length(nf_lambda_check$omega_preds) == ds$tau())

stopifnot(all(vapply(nf_lambda_check$lambda_obs_preds, length, integer(1)) == ds$n))
stopifnot(all(vapply(nf_lambda_check$r_obs_preds, length, integer(1)) == ds$n))
stopifnot(all(vapply(nf_lambda_check$r_d_preds, length, integer(1)) == ds$n))
stopifnot(all(vapply(nf_lambda_check$omega_preds, length, integer(1)) == ds$n))
```


# Fit the LMTP TMLE

```{r}
# TMLE with lambda-learned ratios -----------------------------------------

nf_lambda <- LMTPNuisanceFactory$new(
  policy_seq = policy_shift,
  m_learners = m_learners,
  lambda_learners = lambda_learners,
  clip_lambda_probability = 1e-6,
  truncate_ratio = c(1e-6, 1e6)
)

fluctuation <- LMTPFluctuationSubmodel$new(
  family = stats::gaussian(),
  use_intercept = TRUE
)

fit_lambda <- run_tmle_for_LMTP(
  ds = ds,
  nuisance_factory = nf_lambda,
  fluctuation = fluctuation
)

fit_lambda
```


# Compare against using density estimation directly 

For a simple check, we compare the ratio-classification fit with the density-based fit using 
`nadir::lnr_glm_density()`.

```{r}
if (requireNamespace("nadir", quietly = TRUE)) {
  g_learners <- make_nadir_density_g_learners(
    tau = ds$tau(),
    learner = nadir::lnr_glm_density
  )

  nf_density <- LMTPNuisanceFactory$new(
    policy_seq = policy_shift,
    m_learners = m_learners,
    g_learners = g_learners,
    truncate_density = 1e-12,
    truncate_ratio = c(1e-6, 1e6)
  )

  fit_density <- run_tmle_for_LMTP(
    ds = ds,
    nuisance_factory = nf_density,
    fluctuation = fluctuation
  )

  comparison <- data.frame(
    ratio_strategy = c("classification", "density"),
    estimate = c(fit_lambda$estimate, fit_density$estimate),
    se = c(fit_lambda$se, fit_density$se)
  )

  comparison
}
```

