---
title: "Simulation of LMTPs"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(lmtpi)
library(tidyverse)
library(magrittr)
```

## Scenario: Simple Net-Benefit Effect, No Time-Varying Confounder/Exposure Feedback

Let's imagine as simple a scenario we can: 

  * There is a drug we are studying for its effect on some disease state 
  * People's adherence to the drug daily is $\mathrm{Bernoulli}(\lambda)$ on each day
  and each time-step represents a week, so we 
  * There are time-varying covariates $L_{i,t}$ representing disease state, but 
  that's it.  We will assume if they take their drug, that reduces their disease state. 
  Everyone in the study starts out with a fairly high disease state. 
  * We will introduce the policy up-front so that we can model both factual 
  and counterfactual outcomes in the data generating process.
  
### Modified Treatment Policy Definition 

Imagine that we could get everyone to improve their adherence to the prescribed 
regimen by *2 days (if feasible)* by some kind of encouragement design. 
Obviously that wouldn't encourage people to take the drug more than 
they're supposed to (once every day of the week).

We'd like to know how much of an impact this might have on their outcomes. 

```{r}
policy <- MTP$new(
  smooth_invertible_regions = list(
    function(A, L) { A %btn[]% c(0, 5) },
    function(A, L) { A == 6 },
    function(A, L) { A == 7 }),
  policy = list(
    \(A, L) { A + 2 },
    \(A, L) { A + 1 },
    \(A, L) { A }
  ),
  inverse_policy = list(
    \(A, L) { A - 2 },
    \(A, L) { A - 1 },
    \(A, L) { A }
  ),
  derivative_of_policy = list(
    \(A, L) 1,
    \(A, L) 1,
    \(A, L) 1
  )
)
```

### Data Generating Process

```{r}
# hyperparameters for study 
lambda <- 0.7
n_timesteps <- 4L
n_individuals <- 40L

# generate some data
dgp_simple_scenario <- function(n, t) {
  df <- list()
  
  stopifnot(is.integer(t))
  
  for (i in 1:(t+1)) {
    if (i == 1) {
      newdata <- 
        tibble::tibble(
          id = 1:n,
          t = 1,
          L = rnorm(n = n, mean = 40, sd = 5),
          A = rbinom(n = n, size = 7, prob = lambda)
        )
    } else if (i > 1 && i < t+1) {
      prior_timestep <- 
        df |> dplyr::filter(t == i - 1)
      
      newdata <- 
        prior_timestep |> 
        dplyr::mutate(
          t = i,
          L = L + rnorm(n = 40, mean = 1, sd = 1)*ifelse(A>3, 1, 0) - 
             rnorm(n = 40, mean = 4, sd = 1)*ifelse(A <= 3, 1, 0),
          A = rbinom(n = n, size = 7, prob = (0:7)/7)
        )
    } else if (i == t+1) {
      prior_timestep <- 
        df |> dplyr::filter(t == i - 1)
      
      newdata <- 
        prior_timestep |> 
        dplyr::mutate(
          t = i,
          L = L + rnorm(n = 40, mean = 1, sd = 1)*ifelse(A>3, 1, 0) - 
             rnorm(n = 40, mean = 4, sd = 1)*ifelse(A <= 3, 1, 0),
          A = NA
        )
    }
    df <- dplyr::bind_rows(df, newdata)
  }
  return(df)
}

df <- dgp_simple_scenario(n = n_individuals, t = n_timesteps)

ggplot(df, aes(x = t, y = L, group = id, color = A)) + 
  geom_point() + 
  geom_line(alpha = 0.5, size = 1) + 
  colorspace::scale_color_continuous_sequential(palette = "Plasma", 
                                                rev = FALSE, end = .85) + 
  theme_bw() + 
  labs(
    x = "Time",
    y = "Disease State",
    color = "Doses Taken",
    shape = "Doses Taken",
    title = "Patient History in Simulation Study",
    subtitle = paste0("n = ", n_individuals, ", ", n_timesteps, " time-steps")) + 
  theme(legend.position = 'bottom',
        panel.grid = element_blank()) 
```



```{r}
# reshaping to a wide format for lmtpi 

df_wide <- df |> 
  tidyr::pivot_wider(
    id_cols = 'id',
    names_from = 't',
    values_from = c('L', 'A'))

# name outcome properly and 
# drop extraneous treatment variable for A_5 (NAs from pivoting)
df_wide %<>% rename(Y = L_5)
df_wide %<>% select(-A_5)

lmtp_data <- LMTP_Data_Struct$new(
  data = df_wide,
  id_col = 'id',
  n_timesteps = n_timesteps,
  exposure_cols = paste0('A_', 1:4),
  time_varying_covariate_cols = lapply(1:n_timesteps, \(t) { paste0('L_', t) }),
  outcome_col = 'Y'
)
```

## outcome regression 

To quote the text of the Díaz et al 2023 paper: 

> In particular, $m_t : t = 1, ..., \tau$ may be estimated as follows. Start by
running a preferred regression algorithm of $m_{\tau+1, i} = Y_i$ on
$(A_{\tau,i}, H_{\tau, i})$. Then evaluate the estimator $\hat m_{\tau}$ at
$(A_{\tau, i}^{\mathbb{d}}, H_{\tau, i})$, that is, compute the prediction $\hat
m_{\tau}(A_{\tau, i}^{\mathbb d}, H_{\tau, i})$. Use this prediction as the
pseudo-outcome in a regression on $(A_{\tau-1,i}, H_{\tau, i})$ to obtain an
estimate of $\hat m_{\tau-1}$. Compute the pseudo-outcome $\hat
m_{\tau-1}(A_{\tau-1,i}, H_{\tau, i})$ and iterate the process until obtaining
an estimate for $\hat m_1$.

```{r}
for (timestep_t in (n_timesteps):1) {
  
  # get history up to time t 
  Ht = lmtp_data$H(timestep_t)
  Y_col <- lmtp_data$outcome_col
  
  if (timestep_t == n_timesteps) {
    
    training_dataset <- 
      dplyr::bind_cols(
           Y = lmtp_data$Y,
           A = lmtp_data$A(timestep_t),
           Ht)
    
    A_col = lmtp_data$exposure_cols[timestep_t]
    colnames(training_dataset)[1] <- Y_col
    colnames(training_dataset)[2] <- A_col
    
    model_t <- 
      lm(formula = paste0(Y_col, " ~ ."), data = training_dataset)
    
    mtp_shifted_A_t <- policy$apply_policy(A = lmtp_data$A(timestep_t), L = lmtp_data$L(timestep_t)) # doesn't depend on L
    
    prediction_dataset_under_policy_shift <- dplyr::bind_cols(mtp_shifted_A_t, Ht)
    colnames(prediction_dataset_under_policy_shift)[1] <- A_col
    pseudo_outcome <- predict(model_t, newdata = prediction_dataset_under_policy_shift)
  } else if (timestep_t >= 1) {
    
      training_dataset <- 
      dplyr::bind_cols(
           Y = pseudo_outcome,
           A = lmtp_data$A(timestep_t),
           Ht)
    
      A_col = lmtp_data$exposure_cols[timestep_t]
      colnames(training_dataset)[1] <- Y_col
      colnames(training_dataset)[2] <- A_col
      
      model_t <- 
        lm(formula = paste0(Y_col, " ~ ."), data = training_dataset)
    
      mtp_shifted_A_t <- policy$apply_policy(A = lmtp_data$A(timestep_t), L = lmtp_data$L(timestep_t)) # doesn't depend on L
      
      prediction_dataset_under_policy_shift <- dplyr::bind_cols(mtp_shifted_A_t, Ht)
      colnames(prediction_dataset_under_policy_shift)[1] <- A_col
      pseudo_outcome <- predict(model_t, newdata = prediction_dataset_under_policy_shift)
  }
}
```

## ipw estimate

## doubly robust estimation

```{r}
lmtp_double_robust_estimator <- function(data, mtp, time, ...) {
  # fitting treatment models 
}
```
