---
title: "Small Simulation Study"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(modulatR)
library(TargetedLearning)
```


## A small simulation study

To check that the pipeline is behaving sensibly, we run a small simulation study
for the contrast

\[
E[Y^d] - E[Y].
\]

For the DGP in this vignette, the true contrast is 2.49.

```{r}
# to determine truth: 
n <- 1e6
W <- rnorm(n)
eL1 <- rnorm(n)
eA1 <- rnorm(n)
eL2 <- rnorm(n)
eA2 <- rnorm(n)
eY  <- rnorm(n)

L1 <- 0.5 * W + eL1
A1 <- 0.6 * W + 0.8 * L1 + eA1
L2 <- 0.4 * W + 0.7 * A1 + 0.6 * L1 + eL2
A2 <- 0.3 * W + 0.5 * L2 + eA2
Y  <- 1 + 0.5 * W + 0.4 * L1 + 0.3 * L2 + 1.2 * A1 + 0.8 * A2 + eY

A1d <- A1 + 1
L2d <- 0.4 * W + 0.7 * A1d + 0.6 * L1 + eL2
A2d <- (0.3 * W + 0.5 * L2d + eA2) + 1
Yd  <- 1 + 0.5 * W + 0.4 * L1 + 0.3 * L2d + 1.2 * A1d + 0.8 * A2d + eY

truth_EY <- mean(Y)
truth_EYd <- mean(Yd)
truth_diff <- mean(Yd - Y)

c(EY = truth_EY, EYd = truth_EYd, diff = truth_diff)
```


```{r}
one_rep <- function(n = 2000, delta = 1) {
  W <- rnorm(n)
  L1 <- 0.5 * W + rnorm(n)
  A1 <- 0.6 * W + 0.8 * L1 + rnorm(n)
  L2 <- 0.4 * W + 0.7 * A1 + 0.6 * L1 + rnorm(n)
  A2 <- 0.3 * W + 0.5 * L2 + rnorm(n)
  Y <- 1 + 0.5 * W + 0.4 * L1 + 0.3 * L2 + 1.2 * A1 + 0.8 * A2 + rnorm(n)

  df <- data.frame(
    W = W,
    L1 = L1,
    A1 = A1,
    L2 = L2,
    A2 = A2,
    Y = Y
  )

  ds <- LMTPData$new(
    data = df,
    A_cols = c("A1", "A2"),
    L_cols = list("L1", "L2"),
    W_cols = "W",
    Y_col = "Y"
  )

  policy_seq <- repeat_policy_over_time(
    mtp_additive_shift(delta = delta),
    tau = ds$tau()
  )

  identity_seq <- identity_policy(
    A_type = "continuous",
    tau = ds$tau()
  )

  nf_shift <- LMTPNuisanceFactory$new(
    learners_g = nadir::lnr_glm_density,
    policy_seq = policy_seq,
    A_type = "continuous",
    g_mode = "density",
    cross_fit = FALSE,
    V = 2
  )
  
  nf_identity <- LMTPNuisanceFactory$new(
    learners_g = nadir::lnr_glm_density,
    policy_seq = identity_seq,
    A_type = "continuous",
    g_mode = "density",
    cross_fit = FALSE,
    V = 2
  )
  
  fluct <- LMTPFluctuationSubmodel$new(
    family = stats::gaussian()
  )

  fit_shift <- make_lmtp_tl_task(
    ds = ds,
    policy_seq = policy_seq,
    nuisance_factory = nf_shift,
    fluctuation = fluct,
    learners_Q = stats::glm
  )$run()

  fit_identity <- make_lmtp_tl_task(
    ds = ds,
    policy_seq = identity_seq,
    nuisance_factory = nf_identity,
    fluctuation = fluct,
    learners_Q = stats::glm
  )$run()

  fit_diff <- lmtp_delta_contrast(
    fit1 = fit_shift,
    fit0 = fit_identity,
    transform = "difference"
  )

  c(
    estimate = fit_diff$estimate,
    se = fit_diff$se,
    lower = fit_diff$ci["lower"],
    upper = fit_diff$ci["upper"],
    covered = as.numeric(
      fit_diff$ci["lower"] <= truth_diff && truth_diff <= fit_diff$ci["upper"]
    )
  )
}
```

```{r}
B <- 200
sim_res <- suppressMessages(replicate(B, one_rep(n = 2000, delta = 1)))
sim_res <- t(sim_res)
sim_res <- as.data.frame(sim_res)

summary_df <- data.frame(
  quantity = c("Monte Carlo mean", "Empirical SD", "Mean estimated SE", "Coverage"),
  value = c(
    mean(sim_res$estimate),
    stats::sd(sim_res$estimate),
    mean(sim_res$se),
    mean(sim_res$covered)
  )
)

summary_df
```

```{r}
B <- 300
sim_res <- suppressMessages(replicate(B, one_rep(n = 100, delta = 1)))
sim_res <- t(sim_res)
sim_res <- as.data.frame(sim_res)

summary_df <- data.frame(
  quantity = c("Monte Carlo mean", "Empirical SD", "Mean estimated SE", "Coverage"),
  value = c(
    mean(sim_res$estimate),
    stats::sd(sim_res$estimate),
    mean(sim_res$se),
    mean(sim_res$covered)
  )
)

summary_df
```

```{r}
B <- 300
sim_res <- suppressMessages(replicate(B, one_rep(n = 200, delta = 1)))
sim_res <- t(sim_res)
sim_res <- as.data.frame(sim_res)

summary_df <- data.frame(
  quantity = c("Monte Carlo mean", "Empirical SD", "Mean estimated SE", "Coverage"),
  value = c(
    mean(sim_res$estimate),
    stats::sd(sim_res$estimate),
    mean(sim_res$se),
    mean(sim_res$covered)
  )
)

summary_df
```

```{r}
# data.frame(
#   truth = truth_diff,
#   mc_mean = mean(sim_res$estimate),
#   bias = mean(sim_res$estimate) - truth_diff,
#   empirical_sd = stats::sd(sim_res$estimate),
#   mean_estimated_se = mean(sim_res$se),
#   coverage_95 = mean(sim_res$covered)
# )
```


## Using small super learners with `{nadir}`

The previous section used simple GLMs for the outcome regressions and a Gaussian
density learner for the treatment mechanism. In this section, we use small
Super Learners built with `{nadir}`.

For illustration, we use the learner library

- `nadir::lnr_glm`
- `nadir::lnr_earth`
- `nadir::lnr_ranger`

for the outcome regressions, and a small density-learning library for the
continuous treatment mechanism.

```{r}
library(nadir)

make_example_objects <- function(n = 2000, delta = 1) {
  W <- rnorm(n)
  L1 <- 0.5 * W + rnorm(n)
  A1 <- 0.6 * W + 0.8 * L1 + rnorm(n)
  L2 <- 0.4 * W + 0.7 * A1 + 0.6 * L1 + rnorm(n)
  A2 <- 0.3 * W + 0.5 * L2 + rnorm(n)
  Y <- 1 + 0.5 * W + 0.4 * L1 + 0.3 * L2 + 1.2 * A1 + 0.8 * A2 + rnorm(n)

  df <- data.frame(
    W = W,
    L1 = L1,
    A1 = A1,
    L2 = L2,
    A2 = A2,
    Y = Y
  )

  ds <- LMTPData$new(
    data = df,
    A_cols = c("A1", "A2"),
    L_cols = list("L1", "L2"),
    W_cols = "W",
    Y_col = "Y"
  )

  policy_seq <- repeat_policy_over_time(
    mtp_additive_shift(delta = delta),
    tau = ds$tau()
  )

  identity_seq <- identity_policy(
    A_type = "continuous",
    tau = ds$tau()
  )

  list(
    ds = ds,
    policy_seq = policy_seq,
    identity_seq = identity_seq
  )
}

q_library <- list(
  nadir::lnr_glm,
  nadir::lnr_earth,
  nadir::lnr_ranger
)

g_library <- list(
  nadir::lnr_glm_density
)

objs <- make_example_objects(n = 2000, delta = 1)

ds <- objs$ds
policy_seq <- objs$policy_seq
identity_seq <- objs$identity_seq

nf_shift <- LMTPNuisanceFactory$new(
  learners_g = g_library,
  policy_seq = policy_seq,
  A_type = "continuous",
  g_mode = "density"
)

nf_identity <- LMTPNuisanceFactory$new(
  learners_g = g_library,
  policy_seq = identity_seq,
  A_type = "continuous",
  g_mode = "density"
)

fluct <- LMTPFluctuationSubmodel$new(
  family = stats::gaussian()
)

task_shift_sl <- make_lmtp_tl_task(
  ds = ds,
  policy_seq = policy_seq,
  nuisance_factory = nf_shift,
  fluctuation = fluct,
  learners_Q = q_library
)

task_identity_sl <- make_lmtp_tl_task(
  ds = ds,
  policy_seq = identity_seq,
  nuisance_factory = nf_identity,
  fluctuation = fluct,
  learners_Q = q_library
)

fit_shift_sl <- task_shift_sl$run()
fit_identity_sl <- task_identity_sl$run()

print(fit_shift_sl)
print(fit_identity_sl)

fit_diff_sl <- lmtp_delta_contrast(
  fit1 = fit_shift_sl,
  fit0 = fit_identity_sl,
  transform = "difference"
)

print(fit_diff_sl)

comparison_sl <- data.frame(
  parameter = c("E[Y]", "E[Y^d]", "E[Y^d] - E[Y]"),
  truth = c(truth_EY, truth_EYd, truth_diff),
  estimate = c(
    fit_identity_sl$estimate,
    fit_shift_sl$estimate,
    fit_diff_sl$estimate
  ),
  se = c(
    fit_identity_sl$se,
    fit_shift_sl$se,
    fit_diff_sl$se
  )
)

comparison_sl
```


## Now Simulation Study with `nadir::super_learner()`

```{r}
one_rep_compare <- function(n = 2000,
                            delta = 1,
                            q_learners = stats::glm) {
  W <- rnorm(n)
  L1 <- 0.5 * W + rnorm(n)
  A1 <- 0.6 * W + 0.8 * L1 + rnorm(n)
  L2 <- 0.4 * W + 0.7 * A1 + 0.6 * L1 + rnorm(n)
  A2 <- 0.3 * W + 0.5 * L2 + rnorm(n)
  Y <- 1 + 0.5 * W + 0.4 * L1 + 0.3 * L2 + 1.2 * A1 + 0.8 * A2 + rnorm(n)

  df <- data.frame(
    W = W,
    L1 = L1,
    A1 = A1,
    L2 = L2,
    A2 = A2,
    Y = Y
  )

  ds <- LMTPData$new(
    data = df,
    A_cols = c("A1", "A2"),
    L_cols = list("L1", "L2"),
    W_cols = "W",
    Y_col = "Y"
  )

  policy_seq <- repeat_policy_over_time(
    mtp_additive_shift(delta = delta),
    tau = ds$tau()
  )

  identity_seq <- identity_policy(
    A_type = "continuous",
    tau = ds$tau()
  )

  nf_shift <- LMTPNuisanceFactory$new(
    learners_g = nadir::lnr_glm_density,
    policy_seq = policy_seq,
    A_type = "continuous",
    g_mode = "density",
    cross_fit = TRUE,
    V = 2
  )

  nf_identity <- LMTPNuisanceFactory$new(
    learners_g = nadir::lnr_glm_density,
    policy_seq = identity_seq,
    A_type = "continuous",
    g_mode = "density",
    cross_fit = TRUE,
    V = 2
  )

  fluct <- LMTPFluctuationSubmodel$new(
    family = stats::gaussian()
  )

  fit_shift <- make_lmtp_tl_task(
    ds = ds,
    policy_seq = policy_seq,
    nuisance_factory = nf_shift,
    fluctuation = fluct,
    learners_Q = q_learners
  )$run()

  fit_identity <- make_lmtp_tl_task(
    ds = ds,
    policy_seq = identity_seq,
    nuisance_factory = nf_identity,
    fluctuation = fluct,
    learners_Q = q_learners
  )$run()

  fit_diff <- lmtp_delta_contrast(
    fit1 = fit_shift,
    fit0 = fit_identity,
    transform = "difference"
  )

  c(
    estimate = fit_diff$estimate,
    se = fit_diff$se,
    lower = fit_diff$ci["lower"],
    upper = fit_diff$ci["upper"],
    covered = as.numeric(
      fit_diff$ci["lower"] <= truth_diff && truth_diff <= fit_diff$ci["upper"]
    )
  )
}
```


```{r}
q_library_small <- list(
  nadir::lnr_glm,
  nadir::lnr_earth,
  nadir::lnr_ranger
)
```

```{r}
set.seed(2026)

B <- 200

sim_glm <- replicate(
  B,
  suppressMessages(one_rep_compare(
    n = 200,
    delta = 1,
    q_learners = stats::glm
  ))
)

sim_sl <- replicate(
  B,
  one_rep_compare(
    n = 200,
    delta = 1,
    q_learners = q_library_small
  )
)

sim_glm <- as.data.frame(t(sim_glm))
sim_sl <- as.data.frame(t(sim_sl))

summarize_sim <- function(res, truth) {
  data.frame(
    mc_mean = mean(res$estimate),
    bias = mean(res$estimate) - truth,
    empirical_sd = stats::sd(res$estimate),
    mean_estimated_se = mean(res$se),
    coverage_95 = mean(res$covered)
  )
}

summary_glm <- summarize_sim(sim_glm, truth = truth_diff)
summary_sl <- summarize_sim(sim_sl, truth = truth_diff)

summary_glm$method <- "GLM"
summary_sl$method <- "Super Learner"

sim_summary <- rbind(summary_glm, summary_sl)
sim_summary <- sim_summary[, c("method", "mc_mean", "bias", "empirical_sd", "mean_estimated_se", "coverage_95")]

sim_summary
```
