---
title: "Stress-testing a multi-valued categorical LMTP"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Stress-testing a multi-valued categorical LMTP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, message=FALSE, warning=FALSE}
library(modulatR)
```

# Goal

This vignette stress-tests `{modulatR}` with a genuinely multi-valued,
longitudinal categorical treatment.

At each of three treatment times, treatment takes one of five ordered values:

```{r treatment-support}
dose_levels <- c(
  "none",
  "low",
  "medium",
  "high",
  "very_high"
)

dose_score <- c(
  none = 0,
  low = 1,
  medium = 2,
  high = 3,
  very_high = 4
)
```

The data-generating process includes three treatment times; two evolving
time-varying covariates (disease severity and treatment toxicity);
history-dependent multinomial treatment assignment; treatment-confounder
feedback; a modified treatment policy that may escalate, maintain, or
de-escalate treatment; and a continuous final outcome.

The purpose is not to mimic a specific clinical setting. It is to construct a
nontrivial categorical LMTP for which we can independently simulate the
counterfactual truth and verify that the estimator recovers it.

# Treatment assignment mechanism

Rather than using a uniform treatment distribution, treatment intensity depends
on current severity, current toxicity, baseline risk, frailty, and the previous
treatment level.

```{r treatment-mechanism}
# Treatment mechanism helpers ----

clip <- function(x, lower, upper) {
  pmin(pmax(x, lower), upper)
}

treatment_probabilities <- function(df, t) {
  severity <- df[[paste0("severity", t)]]
  toxicity <- df[[paste0("toxicity", t)]]

  previous_dose <- if (t == 1L) {
    rep(0, nrow(df))
  } else {
    unname(dose_score[df[[paste0("A", t - 1L)]]])
  }

  preferred_dose <-
    1.10 +
    0.75 * severity -
    0.65 * toxicity +
    0.25 * df$baseline_risk +
    0.30 * df$frailty +
    0.35 * previous_dose

  preferred_dose <- clip(
    preferred_dose,
    lower = 0.10,
    upper = 3.90
  )

  score_values <- unname(dose_score[dose_levels])
  category_bias <- c(0.10, 0.05, 0.00, -0.05, -0.10)

  eta <- vapply(
    score_values,
    function(k) {
      -0.70 * (k - preferred_dose)^2
    },
    numeric(nrow(df))
  )

  eta <- sweep(
    eta,
    MARGIN = 2,
    STATS = category_bias,
    FUN = "+"
  )

  eta <- eta - apply(
    eta,
    MARGIN = 1,
    FUN = max
  )

  exp_eta <- exp(eta)
  probs <- exp_eta / rowSums(exp_eta)

  colnames(probs) <- dose_levels

  probs
}

draw_treatment <- function(df,
                           t,
                           u = stats::runif(nrow(df))) {
  probs <- treatment_probabilities(
    df = df,
    t = t
  )

  cumulative_probs <- t(
    apply(
      probs,
      MARGIN = 1,
      FUN = cumsum
    )
  )

  category_index <- 1L +
    rowSums(
      u > cumulative_probs
    )

  dose_levels[category_index]
}
```

For example:

```{r treatment-probability-example}
example_histories <- data.frame(
  baseline_risk = c(-1, 0, 1),
  frailty = c(0, 0, 1),
  severity1 = c(-0.5, 0.5, 1.5),
  toxicity1 = c(0.0, 0.3, 1.2)
)

round(
  treatment_probabilities(
    example_histories,
    t = 1
  ),
  3
)
```

# Define a clinically adaptive categorical LMTP

At each treatment time:

1. if toxicity is high, decrease treatment by one category;
2. otherwise, if disease severity is high and toxicity is sufficiently low,
   increase treatment by one category;
3. otherwise, leave treatment unchanged.

Treatment is capped at `"very_high"` and floored at `"none"`.

```{r policy-helper}
# Policy helper ----

modify_dose <- function(A,
                        severity,
                        toxicity) {
  score <- unname(dose_score[A])

  deescalate <-
    toxicity > 1.00 &
    score > 0

  escalate <-
    !deescalate &
    severity > 0.50 &
    toxicity < 0.75 &
    score < 4

  modified_score <-
    score -
    as.integer(deescalate) +
    as.integer(escalate)

  dose_levels[modified_score + 1L]
}
```

We define a separate discrete MTP at each time so that each policy explicitly
uses the appropriate current covariates.

```{r define-policy}
# Construct time-specific categorical MTPs ----

mtp_t1 <- mtp_discrete(
  map_fun = function(A, H_df) {
    modify_dose(
      A = A,
      severity = H_df$severity1,
      toxicity = H_df$toxicity1
    )
  },
  support = dose_levels,
  name = "adaptive_dose_t1"
)

mtp_t2 <- mtp_discrete(
  map_fun = function(A, H_df) {
    modify_dose(
      A = A,
      severity = H_df$severity2,
      toxicity = H_df$toxicity2
    )
  },
  support = dose_levels,
  name = "adaptive_dose_t2"
)

mtp_t3 <- mtp_discrete(
  map_fun = function(A, H_df) {
    modify_dose(
      A = A,
      severity = H_df$severity3,
      toxicity = H_df$toxicity3
    )
  },
  support = dose_levels,
  name = "adaptive_dose_t3"
)

policy_adaptive <- LMTPPolicySequence$new(
  policies = list(
    mtp_t1,
    mtp_t2,
    mtp_t3
  ),
  name = "adaptive_escalation_deescalation"
)
```

# Longitudinal data-generating process

Higher treatment intensity tends to reduce subsequent disease severity but
increase subsequent toxicity. Both severity and toxicity affect later treatment
assignment.

```{r exogenous-data}
# Generate exogenous variables ----

generate_exogenous_data <- function(n) {
  data.frame(
    baseline_risk = stats::rnorm(n),
    frailty = stats::rbinom(
      n,
      size = 1,
      prob = 0.35
    ),

    e_severity1 = stats::rnorm(n),
    e_toxicity1 = stats::rnorm(n),

    e_severity2 = stats::rnorm(n),
    e_toxicity2 = stats::rnorm(n),

    e_severity3 = stats::rnorm(n),
    e_toxicity3 = stats::rnorm(n),

    u_A1 = stats::runif(n),
    u_A2 = stats::runif(n),
    u_A3 = stats::runif(n),

    e_Y = stats::rnorm(n)
  )
}
```

The same function can generate either the observed trajectory or the trajectory
under the LMTP. Under the LMTP, later covariates and later natural treatment
draws evolve using the modified prior treatment history.

```{r trajectory-generator}
# Simulate one longitudinal trajectory regime ----

simulate_trajectory <- function(exogenous,
                                apply_policy = FALSE) {
  baseline_risk <- exogenous$baseline_risk
  frailty <- exogenous$frailty

  severity1 <-
    0.70 * baseline_risk +
    0.45 * frailty +
    exogenous$e_severity1

  toxicity1 <-
    0.20 * frailty +
    0.50 * exogenous$e_toxicity1

  H1 <- data.frame(
    baseline_risk = baseline_risk,
    frailty = frailty,
    severity1 = severity1,
    toxicity1 = toxicity1
  )

  A1_natural <- draw_treatment(
    df = H1,
    t = 1,
    u = exogenous$u_A1
  )

  A1 <- if (apply_policy) {
    modify_dose(
      A = A1_natural,
      severity = severity1,
      toxicity = toxicity1
    )
  } else {
    A1_natural
  }

  dose1 <- unname(dose_score[A1])

  severity2 <-
    0.55 * severity1 +
    0.30 * baseline_risk +
    0.20 * frailty -
    0.30 * dose1 +
    0.12 * severity1 * frailty +
    exogenous$e_severity2

  toxicity2 <-
    0.45 * toxicity1 +
    0.20 * frailty +
    0.28 * dose1 +
    0.06 * dose1^2 +
    0.60 * exogenous$e_toxicity2

  H2 <- data.frame(
    baseline_risk = baseline_risk,
    frailty = frailty,
    severity1 = severity1,
    toxicity1 = toxicity1,
    A1 = A1,
    severity2 = severity2,
    toxicity2 = toxicity2
  )

  A2_natural <- draw_treatment(
    df = H2,
    t = 2,
    u = exogenous$u_A2
  )

  A2 <- if (apply_policy) {
    modify_dose(
      A = A2_natural,
      severity = severity2,
      toxicity = toxicity2
    )
  } else {
    A2_natural
  }

  dose2 <- unname(dose_score[A2])

  severity3 <-
    0.50 * severity2 +
    0.20 * baseline_risk +
    0.15 * frailty -
    0.32 * dose2 +
    0.08 * severity2^2 +
    exogenous$e_severity3

  toxicity3 <-
    0.50 * toxicity2 +
    0.15 * frailty +
    0.30 * dose2 +
    0.05 * dose2^2 +
    0.60 * exogenous$e_toxicity3

  H3 <- data.frame(
    baseline_risk = baseline_risk,
    frailty = frailty,
    severity1 = severity1,
    toxicity1 = toxicity1,
    A1 = A1,
    severity2 = severity2,
    toxicity2 = toxicity2,
    A2 = A2,
    severity3 = severity3,
    toxicity3 = toxicity3
  )

  A3_natural <- draw_treatment(
    df = H3,
    t = 3,
    u = exogenous$u_A3
  )

  A3 <- if (apply_policy) {
    modify_dose(
      A = A3_natural,
      severity = severity3,
      toxicity = toxicity3
    )
  } else {
    A3_natural
  }

  dose3 <- unname(dose_score[A3])

  # Higher outcome values are better.
  Y <-
    10 +
    0.30 * baseline_risk -
    0.20 * frailty -
    0.35 * severity2 -
    1.00 * severity3 -
    0.25 * toxicity2 -
    0.75 * toxicity3 +
    0.20 * dose3 -
    0.04 * dose3^2 +
    exogenous$e_Y

  data.frame(
    baseline_risk = baseline_risk,
    frailty = frailty,

    severity1 = severity1,
    toxicity1 = toxicity1,
    A1 = A1,

    severity2 = severity2,
    toxicity2 = toxicity2,
    A2 = A2,

    severity3 = severity3,
    toxicity3 = toxicity3,
    A3 = A3,

    Y = Y,
    stringsAsFactors = FALSE
  )
}
```

# Monte Carlo truth

```{r monte-carlo-truth}
# Monte Carlo truth ----

set.seed(1001)

n_truth <- 250000

exo_truth <- generate_exogenous_data(
  n = n_truth
)

truth_natural_data <- simulate_trajectory(
  exogenous = exo_truth,
  apply_policy = FALSE
)

truth_policy_data <- simulate_trajectory(
  exogenous = exo_truth,
  apply_policy = TRUE
)

truth <- c(
  EY = mean(truth_natural_data$Y),
  EYd = mean(truth_policy_data$Y),
  difference = mean(
    truth_policy_data$Y -
      truth_natural_data$Y
  )
)

truth_mc_se <- c(
  EY = stats::sd(
    truth_natural_data$Y
  ) / sqrt(n_truth),

  EYd = stats::sd(
    truth_policy_data$Y
  ) / sqrt(n_truth),

  difference = stats::sd(
    truth_policy_data$Y -
      truth_natural_data$Y
  ) / sqrt(n_truth)
)

rbind(
  truth = truth,
  Monte_Carlo_SE = truth_mc_se
)
```

# Generate an observed dataset

```{r observed-data}
# Observed estimation sample ----

set.seed(2002)

n <- 15000

observed_data <- simulate_trajectory(
  exogenous = generate_exogenous_data(n),
  apply_policy = FALSE
)

ds <- LMTPData$new(
  data = observed_data,

  W_cols = c(
    "baseline_risk",
    "frailty"
  ),

  L_cols = list(
    c(
      "severity1",
      "toxicity1"
    ),
    c(
      "severity2",
      "toxicity2"
    ),
    c(
      "severity3",
      "toxicity3"
    )
  ),

  A_cols = c(
    "A1",
    "A2",
    "A3"
  ),

  Y_col = "Y"
)

ds
```

The natural treatment distributions are genuinely multi-valued at every time:

```{r observed-treatment-distributions}
# Observed treatment distributions ----

prop.table(table(observed_data$A1))
prop.table(table(observed_data$A2))
prop.table(table(observed_data$A3))
```

# Inspect how the policy modifies treatment

```{r inspect-policy}
# Inspect policy modifications ----

A1_d <- policy_adaptive$apply_t(
  t = 1,
  A_vec = ds$A(1),
  H_df = ds$H(1)
)

A2_d <- policy_adaptive$apply_t(
  t = 2,
  A_vec = ds$A(2),
  H_df = ds$H(2)
)

A3_d <- policy_adaptive$apply_t(
  t = 3,
  A_vec = ds$A(3),
  H_df = ds$H(3)
)

table(
  observed = ds$A(1),
  modified = A1_d
)

table(
  observed = ds$A(2),
  modified = A2_d
)

table(
  observed = ds$A(3),
  modified = A3_d
)
```

```{r policy-change-frequency}
# Policy-change frequencies ----

data.frame(
  time = 1:3,

  proportion_changed = c(
    mean(ds$A(1) != A1_d),
    mean(ds$A(2) != A2_d),
    mean(ds$A(3) != A3_d)
  )
)
```

# Oracle categorical treatment-mass learners

For this stress test, we use the known treatment assignment mechanism rather
than estimating it. This isolates the discrete LMTP machinery: evaluation of
\(g_t(a \mid h)\), construction of \(g_t^d(a \mid h)\), longitudinal density
ratios, and the targeting recursion.

```{r oracle-g-learners}
# Oracle categorical g learners ----

make_oracle_g_learner <- function(t) {
  force(t)

  function(data) {
    function(newdata) {
      probs <- treatment_probabilities(
        df = newdata$df,
        t = t
      )

      A_col <- paste0(
        "A",
        t
      )

      treatment_index <- match(
        newdata$df[[A_col]],
        dose_levels
      )

      if (anyNA(treatment_index)) {
        stop(
          "Encountered a treatment value outside `dose_levels`."
        )
      }

      probs[
        cbind(
          seq_len(newdata$n),
          treatment_index
        )
      ]
    }
  }
}

g_learners <- lapply(
  1:3,
  make_oracle_g_learner
)
```

# Outcome-regression learners

The outcome regressions are estimated rather than supplied from the truth.

```{r m-learners}
# Sequential outcome-regression learners ----

m_formula_t1 <-
  `..pseudo_outcome` ~
  baseline_risk +
  I(baseline_risk^2) +
  frailty +
  severity1 +
  I(severity1^2) +
  toxicity1 +
  I(toxicity1^2) +
  A1 +
  severity1:toxicity1

m_formula_t2 <-
  `..pseudo_outcome` ~
  baseline_risk +
  I(baseline_risk^2) +
  frailty +
  severity1 +
  toxicity1 +
  A1 +
  severity2 +
  I(severity2^2) +
  toxicity2 +
  I(toxicity2^2) +
  A2 +
  severity2:toxicity2

m_formula_t3 <-
  `..pseudo_outcome` ~
  baseline_risk +
  I(baseline_risk^2) +
  frailty +
  severity1 +
  toxicity1 +
  A1 +
  severity2 +
  toxicity2 +
  A2 +
  severity3 +
  I(severity3^2) +
  toxicity3 +
  I(toxicity3^2) +
  A3 +
  severity3:toxicity3

m_learners <- make_glm_m_learners(
  tau = ds$tau(),
  family = stats::gaussian(),
  formula = list(
    m_formula_t1,
    m_formula_t2,
    m_formula_t3
  )
)
```

# Fit the LMTP TMLE

```{r fit-tmle}
# Fit adaptive categorical LMTP TMLE ----

nuisance_factory <- LMTPNuisanceFactory$new(
  policy_seq = policy_adaptive,
  m_learners = m_learners,
  g_learners = g_learners,
  truncate_density = 1e-12,
  truncate_ratio = c(
    1e-6,
    1e6
  )
)

fluctuation <- LMTPFluctuationSubmodel$new(
  family = stats::gaussian()
)

fit <- run_tmle_for_LMTP(
  ds = ds,
  nuisance_factory = nuisance_factory,
  fluctuation = fluctuation
)

fit
```

# Compare the estimate with the counterfactual truth

```{r compare-truth}
# Compare estimate with Monte Carlo truth ----

comparison <- data.frame(
  estimand = "E[Y^d]",
  truth = unname(
    truth["EYd"]
  ),
  truth_mc_se = unname(
    truth_mc_se["EYd"]
  ),
  estimate = unname(
    fit$estimate
  ),
  se = unname(
    fit$se
  ),
  lower = unname(
    fit$ci["lower"]
  ),
  upper = unname(
    fit$ci["upper"]
  )
)

comparison$error <-
  comparison$estimate -
  comparison$truth

comparison$z_from_truth <-
  comparison$error /
  sqrt(
    comparison$se^2 +
      comparison$truth_mc_se^2
  )

comparison
```

For an executable end-to-end check, require the estimate to lie within four
combined standard errors of the independently simulated truth.

```{r executable-check}
# End-to-end sanity check ----

combined_se <- sqrt(
  fit$se^2 +
    truth_mc_se["EYd"]^2
)

stopifnot(
  abs(
    fit$estimate -
      truth["EYd"]
  ) <
    4 * combined_se
)
```

# Inspect categorical density-ratio diagnostics

```{r ratio-diagnostics}
# Density-ratio diagnostics ----

ratio_diagnostics <- data.frame(
  time = 1:ds$tau(),

  mean_r = vapply(
    nuisance_factory$r_obs_preds,
    mean,
    numeric(1)
  ),

  min_r = vapply(
    nuisance_factory$r_obs_preds,
    min,
    numeric(1)
  ),

  max_r = vapply(
    nuisance_factory$r_obs_preds,
    max,
    numeric(1)
  ),

  mean_omega = vapply(
    nuisance_factory$omega_preds,
    mean,
    numeric(1)
  ),

  min_omega = vapply(
    nuisance_factory$omega_preds,
    min,
    numeric(1)
  ),

  max_omega = vapply(
    nuisance_factory$omega_preds,
    max,
    numeric(1)
  )
)

ratio_diagnostics
```

# What this vignette verifies

This example exercises substantially more of the categorical-treatment path
than a simple binary or uniformly distributed exposure. It checks an LMTP with
five treatment categories, three longitudinal decisions, history-dependent
multinomial assignment, treatment-confounder feedback, multiple-to-one policy
mappings, both escalation and de-escalation, exact discrete pushforward masses,
longitudinal density-ratio products, sequential regression, targeting, and
comparison against an independently generated Monte Carlo counterfactual truth.

If this vignette runs successfully and the final estimate agrees with the Monte
Carlo truth, that is strong evidence that the main density-based categorical
LMTP pathway in `{modulatR}` is functioning correctly.

It does not separately validate the observed-versus-shifted classification
ratio route. That should be tested separately so failures in that nuisance
strategy are not conflated with failures of the discrete MTP machinery itself.
