---
title: "Subgroup TMLE for Longitudinal Modified Treatment Policies"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(modulatR)
library(nadir)
library(TargetedLearning)
```


# Define a dataset

This dataset is constructed to have effect modification according to the baseline
variable $W$. The DGP written below (`generate_lmtp_data()`) returns not 
only a dataset `(W, L1, A1, L2, A2, Y)` but also the column `Ystar` which is their
counterfactual outcomes under a policy of an additive shift $d(A) = A+1$ at both
time-steps.

```{r}
generate_lmtp_data <- function(n, delta = 1) {
  W  <- rnorm(n, sd = 10)

  eL1 <- rnorm(n)
  eA1 <- rnorm(n)
  eL2 <- rnorm(n)
  eA2 <- rnorm(n)
  eY  <- rnorm(n, sd = 5)

  L1 <- 0.5 * W + eL1
  A1 <- 0.6 * W + 0.8 * L1 + eA1
  L2 <- 0.4 * W + 0.7 * A1 + 0.6 * L1 + eL2
  A2 <- 0.3 * W + 0.5 * L2 + eA2

  Y <- 1 +
    0.5 * W +
    0.4 * L1 +
    0.3 * L2 +
    1.2 * A1 +
    0.8 * A2 +
    0.15 * W * A1 +
    0.10 * W * A2 +
    eY

  A1_star <- A1 + delta
  L2_star <- 0.4 * W + 0.7 * A1_star + 0.6 * L1 + eL2
  A2_star <- 0.3 * W + 0.5 * L2_star + eA2 + delta

  Y_star <- 1 +
    0.5 * W +
    0.4 * L1 +
    0.3 * L2_star +
    1.2 * A1_star +
    0.8 * A2_star +
    0.15 * W * A1_star +
    0.10 * W * A2_star +
    eY

  data.frame(
    W = W,
    L1 = L1,
    A1 = A1,
    L2 = L2,
    A2 = A2,
    Y = Y,
    Y_star = Y_star
  )
}
```

# Approximate truth under dgp 

Here our purpose is to determine and save the approximate counterfactuals
`Ystar` so that we know if our `{modulatR}` software is working correctly or not.

```{r}
set.seed(2026)
truth_dat <- generate_lmtp_data(1e6, delta = 1)

truth_overall <- mean(truth_dat$Y_star)
truth_w_pos   <- mean(truth_dat$Y_star[truth_dat$W > 0])
truth_w_nonpos <- mean(truth_dat$Y_star[truth_dat$W <= 0])

truth_tbl <- data.frame(
  subgroup = c("Overall", "W > 0", "W <= 0"),
  truth = c(truth_overall, truth_w_pos, truth_w_nonpos)
)

truth_tbl
```

# Simulation dataset

```{r}
dat <- generate_lmtp_data(1500, delta = 1)
dat <- dat[, c("W", "L1", "A1", "L2", "A2", "Y")]
head(dat)
```


# Encode data as `LMTPData`
```{r}
ds <- LMTPData$new(
  data   = dat,
  W_cols = "W",
  L_cols = list("L1", "L2"),
  A_cols = c("A1", "A2"),
  Y_col  = "Y"
)

ds
```

# Define a policy 

```{r}
pol <- mtp_additive_shift(delta = 1)
policy_seq <- repeat_policy_over_time(pol, tau = ds$tau())
policy_seq
```

# Specify nuisance estimation and fluctuation

```{r}
nf <- LMTPNuisanceFactory$new(
  learners_g = nadir::lnr_glm_density,
  policy_seq = policy_seq,
  A_type = "continuous",
  g_mode = "density"
)

fluct <- LMTPFluctuationSubmodel$new(
  family = stats::gaussian()
)
```

# Define subgroups

```{r}
subgroup_funs <- list(
  W_gt_0   = function(df) as.numeric(df$W > 0),
  W_leq_0  = function(df) as.numeric(df$W <= 0)
)
```

# Run subgroup TMLE

```{r}
fit_subgroup <- run_subgroup_tmle_for_LMTP(
  ds = ds,
  policy_seq = policy_seq,
  nuisance_factory = nf,
  fluctuation = fluct,
  subgroup_funs = subgroup_funs,
  learners_Q = nadir::lnr_glm
)

fit_subgroup
```

# Inspect estimates

```{r}
fit_subgroup$estimate

# ci
fit_subgroup$se
fit_subgroup$ci

# eif
dim(fit_subgroup$eif)
colnames(fit_subgroup$eif)
head(fit_subgroup$eif)
```

# Compare to truth

```{r}
est_tbl <- data.frame(
  subgroup = names(fit_subgroup$estimate),
  estimate = as.numeric(fit_subgroup$estimate),
  se = as.numeric(fit_subgroup$se),
  ci_lower = fit_subgroup$ci["lower", ],
  ci_upper = fit_subgroup$ci["upper", ]
)

est_tbl$subgroup <- c(`W_gt_0` = 'W > 0', `W_leq_0` = 'W <= 0')[est_tbl$subgroup]

merge(est_tbl, truth_tbl, by = "subgroup", all.x = TRUE)
```

