---
title: "Estimating subgroup-specific LMTP means"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimating subgroup-specific LMTP means}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)

set.seed(2026)

library(modulatR)
library(nadir)
```


This vignette illustrates the simultaneous subgroup TMLE implemented in
`run_subgroup_tmle_for_LMTP()`.

The target parameters are subgroup-specific LMTP means,
$$\psi_j = \mathbb E\{Y(\bar A^d) \mid V \in \mathcal A_j\},
\quad j = 1,\ldots,J.$$

The function targets all subgroup parameters simultaneously using a vector-valued
clever covariate. This follows the same broad idea as multiple-subgroup targeted
learning: rather than targeting each subgroup one at a time, we construct a
matrix of subgroup-scaled clever covariates and update the outcome regression in
one joint targeting step. Wei et al. discuss the motivation for simultaneous
multi-subgroup targeting and also propose stabilized refinements for finite
sample settings; here we use the simpler direct vector-valued targeting approach.

## Truth by Monte Carlo Sample

```{r}
# Truth by large Monte Carlo sample ----------------------------------------

set.seed(1)

n_truth <- 3e5

W   <- rnorm(n_truth)
eL1 <- rnorm(n_truth)
eA1 <- rnorm(n_truth)
eL2 <- rnorm(n_truth)
eA2 <- rnorm(n_truth)
eY  <- rnorm(n_truth)

L1 <- 0.5 * W + eL1
A1 <- 0.6 * W + 0.8 * L1 + eA1
L2 <- 0.4 * W + 0.7 * A1 + 0.6 * L1 + eL2
A2 <- 0.3 * W + 0.5 * L2 + eA2

Y <- 1 +
  0.5 * W +
  0.4 * L1 +
  0.3 * L2 +
  1.2 * A1 +
  0.8 * A2 +
  eY

A1d <- A1 + 1
L2d <- 0.4 * W + 0.7 * A1d + 0.6 * L1 + eL2
A2d <- 0.3 * W + 0.5 * L2d + eA2 + 1

Yd <- 1 +
  0.5 * W +
  0.4 * L1 +
  0.3 * L2d +
  1.2 * A1d +
  0.8 * A2d +
  eY

truth_df <- data.frame(W, L1, A1, L2, A2, Y, Yd)

truth_subgroups <- list(
  W_positive = function(df) as.numeric(df$W > 0),
  L1_positive = function(df) as.numeric(df$L1 > 0),
  high_baseline = function(df) as.numeric(df$W > 0 & df$L1 > 0)
)

truth_subgroup_means <- vapply(
  truth_subgroups,
  function(g) {
    idx <- g(truth_df) == 1
    mean(truth_df$Yd[idx])
  },
  numeric(1)
)

truth_subgroup_means
```


## Observed Data

```{r}
# Observed data ------------------------------------------------------------

set.seed(2)

n <- 8000

W  <- rnorm(n)
L1 <- 0.5 * W + rnorm(n)
A1 <- 0.6 * W + 0.8 * L1 + rnorm(n)
L2 <- 0.4 * W + 0.7 * A1 + 0.6 * L1 + rnorm(n)
A2 <- 0.3 * W + 0.5 * L2 + rnorm(n)

Y <- 1 +
  0.5 * W +
  0.4 * L1 +
  0.3 * L2 +
  1.2 * A1 +
  0.8 * A2 +
  rnorm(n)

df <- data.frame(W, L1, A1, L2, A2, Y)

ds <- LMTPData$new(
  data = df,
  A_cols = c("A1", "A2"),
  L_cols = list("L1", "L2"),
  W_cols = "W",
  Y_col = "Y"
)

ds
```


## Define policy and subgroups

```{r}
# Policy and subgroups -----------------------------------------------------

policy_shift <- repeat_policy_over_time(
  mtp_additive_shift(delta = 1),
  tau = ds$tau()
)

subgroup_funs <- list(
  W_positive = function(df) as.numeric(df$W > 0),
  L1_positive = function(df) as.numeric(df$L1 > 0),
  high_baseline = function(df) as.numeric(df$W > 0 & df$L1 > 0)
)

subgroup_mat <- modulatR:::.make_subgroup_matrix(ds, subgroup_funs)

colMeans(subgroup_mat)
```

## Nuisance Learners

```{r}
# Nuisance learners --------------------------------------------------------

m_learners <- make_glm_m_learners(
  tau = ds$tau(),
  family = stats::gaussian()
)

g_learners <- make_nadir_density_g_learners(
  tau = ds$tau(),
  learner = nadir::lnr_glm_density
)
```

## Fit Subgroup LMTP TMLE

```{r}
# Nuisance factory and fluctuation -----------------------------------------

nf_shift <- LMTPNuisanceFactory$new(
  policy_seq = policy_shift,
  m_learners = m_learners,
  g_learners = g_learners,
  truncate_density = 1e-12,
  truncate_ratio = c(1e-6, Inf)
)

fluctuation <- LMTPFluctuationSubmodel$new(
  family = stats::gaussian(),
  use_intercept = TRUE
)

# Subgroup TMLE ------------------------------------------------------------

fit_subgroup <- run_subgroup_tmle_for_LMTP(
  ds = ds,
  nuisance_factory = nf_shift,
  subgroup_funs = subgroup_funs,
  fluctuation = fluctuation
)

fit_subgroup
```

## Comparison with Truth

```{r}
# Comparison with Monte Carlo truth ----------------------------------------

comparison <- data.frame(
  subgroup = names(truth_subgroup_means),
  truth = as.numeric(truth_subgroup_means),
  estimate = as.numeric(fit_subgroup$estimate),
  se = as.numeric(fit_subgroup$se),
  lower = as.numeric(fit_subgroup$ci["lower", ]),
  upper = as.numeric(fit_subgroup$ci["upper", ])
)

comparison$error <- comparison$estimate - comparison$truth

comparison
```


## Inspect the EIF Matrix

```{r}
# EIF matrix ---------------------------------------------------------------

dim(fit_subgroup$eif)
colnames(fit_subgroup$eif)

head(fit_subgroup$eif)

apply(fit_subgroup$eif, 2, mean)
apply(fit_subgroup$eif, 2, var)
```

## Check on the stored objects

```{r}
# Stored fitted quantities -------------------------------------------------

names(fit_subgroup$m_init)
names(fit_subgroup$m_star)

length(fit_subgroup$m_init$m_obs)
length(fit_subgroup$m_star$m_d)

str(fit_subgroup$m_init, max.level = 1)
str(fit_subgroup$m_star, max.level = 1)
```


